% Schubert 多项式学习笔记
% 学习文献：
% 1. Graham Positivity of Triple Schubert Calculus (Gao & Xiong, 2025) <- 当前学习
% 2. Molev-Sagan Formula for Double Schubert Polynomials (Samuel, 2024)
% 3. Quantum Double Schubert Polynomials (Kirillov & Maeno, 1996)
% 4. Positivity in Equivariant Quantum Schubert Calculus (Mihalcea)

\documentclass[12pt]{amsbook}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{xeCJK}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{arrows,positioning}
\geometry{margin=1in}

% Theorem styles
\theoremstyle{plain}
\newtheorem{Definition}{定义}[chapter]
\newtheorem{Theorem}[Definition]{定理}
\newtheorem{Lemma}[Definition]{引理}
\newtheorem{Corollary}[Definition]{推论}
\newtheorem{Proposition}[Definition]{命题}
\newtheorem{Example}{例}[chapter]
\newtheorem{Remark}{注}[chapter]
\newtheorem{Conjecture}[Definition]{猜想}

% User annotation command
\newcommand{\userannotation}[2]{%
  \begin{Remark}
    \textbf{#1:} #2
  \end{Remark}
}

\title{Schubert 多项式学习笔记}
\author{}
\date{\today}

\begin{document}

\maketitle
\tableofcontents

\input{chapters/chapter0}
\input{chapters/chapter1}

\appendix
\input{appendix/qa}

\begin{thebibliography}{99}

 bibitem{Gr} W. Graham, \emph{Positivity in equivariant Schubert calculus}, Duke Math. J. 109 (2001), no. 3, 449--477.

 bibitem{GX2025} Y. B. Gao, R. Xiong, \emph{Graham positivity of triple Schubert calculus}, arXiv:2506.09421.

 bibitem{Sa} M. J. Samuel, \emph{A Molev-Sagan type formula for double Schubert polynomials}, arXiv:2401.11060.

 bibitem{Ma} I. G. Macdonald, \emph{Notes on Schubert polynomials}, Publ. LACIM 6, UQAM, 1991.

 bibitem{KM} A. N. Kirillov, T. Maeno, \emph{Quantum double Schubert polynomials}, arXiv:q-alg/9610022.

 bibitem{Buch2001} A. S. Buch, \emph{Quantum cohomology of Grassmannians}, arXiv:math/0106268.

 bibitem{Li2023} C. Z. Li, \emph{Schubert calculus -- A brief introduction}, 2023.

 bibitem{Li2023b} C. Z. Li, \emph{Certain identities on the quantum product of Schubert classes}, 2023.

 bibitem{AndersonFulton} D. Anderson, W. Fulton, \emph{Equivariant Cohomology in Algebraic Geometry}, 2015.

\end{thebibliography}

\end{document}
