\chapter{Distributions of Two Random Variables}\label{chap:distributions-two-rv}

\section{Distributions of Two Random Variables}\label{sec:distributions-two-rv}

We begin the discussion of a pair of random variables with the following example. A coin is tossed three times and our interest is in the ordered number pair (number of H's on first two tosses, number of H's on all three tosses), where H and T represent, respectively, heads and tails. Let $\mathcal{C} = \{TTT,TTH,THT,HTT,THH,HTH,HHT,HHH\}$ denote the sample space. Let $X_{1}$ denote the number of H's on the first two tosses and $X_{2}$ denote the number of H's on all three flips. Then our interest can be represented by the pair of random variables $(X_{1},X_{2})$ . For example, $(X_{1}(HTH),X_{2}(HTH))$ represents the outcome $(1,2)$ . Continuing in this way, $X_{1}$ and $X_{2}$ are real-valued functions defined on the sample space $\mathcal{C}$ , which take us from the sample space to the space of ordered number pairs.

\[
\mathcal{D} = \{(0,0), (0,1), (1,1), (1,2), (2,2), (2,3)\}.
\]

Thus $X_{1}$ and $X_{2}$ are two random variables defined on the space $\mathcal{C}$ , and, in this example, the space of these random variables is the two-dimensional set $\mathcal{D}$ , which is a subset of two-dimensional Euclidean space $R^{2}$ . Hence $(X_{1},X_{2})$ is a vector function from $\mathcal{C}$ to $\mathcal{D}$ . We now formulate the definition of a random vector.

\begin{definition}[Random Vector]\label{def:random-vector}
Given a random experiment with a sample space $\mathcal{C}$ , consider two random variables $X_{1}$ and $X_{2}$ , which assign to each element $c$ of $\mathcal{C}$ one and only one ordered pair of numbers $X_{1}(c) = x_{1}$ , $X_{2}(c) = x_{2}$ . Then we say that $(X_{1},X_{2})$ is a random vector. The space of $(X_{1},X_{2})$ is the set of ordered pairs $\mathcal{D} = \{(x_{1},x_{2}) : x_{1} = X_{1}(c), x_{2} = X_{2}(c), c \in \mathcal{C}\}$ .
\end{definition}

We often denote random vectors using vector notation $\mathbf{X} = (X_{1},X_{2})'$ , where the prime denotes the transpose of the row vector $(X_{1},X_{2})$ . Also, we often use $(X,Y)$ to denote random vectors.

Let $\mathcal{D}$ be the space associated with the random vector $(X_{1},X_{2})$ . Let $A$ be a subset of $\mathcal{D}$ . As in the case of one random variable, we speak of the event $A$ . We wish to define the probability of the event $A$ , which we denote by $P_{X_{1},X_{2}}[A]$ . As with random variables in Section 1.5 we can uniquely define $P_{X_{1},X_{2}}$ in terms of the cumulative distribution function (cdf), which is given by

\begin{equation}\label{eq:joint-cdf}
F_{X_{1},X_{2}}(x_{1},x_{2}) = P[X_{1} \leq x_{1}, X_{2} \leq x_{2}],
\end{equation}

for all $(x_{1},x_{2}) \in R^{2}$ . Because $X_{1}$ and $X_{2}$ are random variables, each of the events in the above intersection and the intersection of the events are events in the original sample space $\mathcal{C}$ . Thus the expression is well defined.

For all real constants $a < b$ , $c < d$ ,

\[
P(a_{1} < X_{1} \leq b_{1}, a_{2} < X_{2} \leq b_{2}) = F(b_{1},b_{2}) - F(a_{1},b_{2}) - F(b_{1},a_{2}) + F(a_{1},a_{2}).
\tag{2.1.2}
\]

Hence, all induced probabilities of sets of the form $(a_{1},b_{1}] \times (a_{2},b_{2}]$ can be formulated in terms of the cdf. We often call this cdf the joint cumulative distribution function of $(X_{1},X_{2})$ .

As with random variables, we are mainly concerned with two types of random vectors, namely discrete and continuous. We first discuss the discrete type.

A random vector $(X_{1},X_{2})$ is a discrete random vector if its space $\mathcal{D}$ is finite or countable. Hence, $X_{1}$ and $X_{2}$ are both discrete also. The joint probability mass function (pmf) of $(X_{1},X_{2})$ is defined by

\begin{equation}\label{eq:joint-pmf}
p_{X_{1},X_{2}}(x_{1},x_{2}) = P[X_{1} = x_{1}, X_{2} = x_{2}],
\end{equation}

for all $(x_{1},x_{2}) \in \mathcal{D}$ . As with random variables, the pmf uniquely defines the cdf. It also is characterized by the two properties

\begin{equation}\label{eq:pmf-properties}
\text{(i) } 0 \leq p_{X_{1},X_{2}}(x_{1},x_{2}) \leq 1 \text{ and (ii)} \sum_{\mathcal{D}} \sum p_{X_{1},X_{2}}(x_{1},x_{2}) = 1.
\end{equation}

For an event $B \in \mathcal{D}$ , we have

\[
P[(X_{1},X_{2}) \in B] = \sum\sum_{B} p_{X_{1},X_{2}}(x_{1},x_{2}).
\]

\begin{example}\label{ex:coin-toss-joint-pmf}
Consider the example at the beginning of this section where a fair coin is flipped three times and $X_{1}$ and $X_{2}$ are the number of heads on the first two flips and all 3 flips, respectively. We can conveniently table the pmf of $(X_{1},X_{2})$ as follows:

\begin{center}
\begin{tabular}{c|cccc|c}
 & $x_{2}=0$ & $x_{2}=1$ & $x_{2}=2$ & $x_{2}=3$ & $p_{X_{1}}(x_{1})$ \\ \hline
$x_{1}=0$ & $1/8$ & $1/8$ & $0$ & $0$ & $2/8$ \\
$x_{1}=1$ & $0$ & $2/8$ & $2/8$ & $0$ & $4/8$ \\
$x_{1}=2$ & $0$ & $0$ & $1/8$ & $1/8$ & $2/8$ \\ \hline
$p_{X_{2}}(x_{2})$ & $1/8$ & $3/8$ & $3/8$ & $1/8$ &
\end{tabular}
\end{center}

For instance, $P(X_{1} \geq 2, X_{2} \geq 2) = p(2,2) + p(2,3) = 2/8$ .
\end{example}

At times it is convenient to speak of the support of a discrete random vector $(X_{1},X_{2})$ . These are all the points $(x_{1},x_{2})$ in the space of $(X_{1},X_{2})$ such that $p(x_{1},x_{2}) > 0$ . In the last example the support consists of the six points $\{(0,0),(0,1),(1,1),(1,2),(2,2),(2,3)\}$ .

We say a random vector $(X_{1},X_{2})$ with space $\mathcal{D}$ is of the continuous type if its cdf $F_{X_{1},X_{2}}(x_{1},x_{2})$ is continuous. For the most part, the continuous random vectors in this book have cdfs that can be represented as integrals of nonnegative functions. That is, $F_{X_{1},X_{2}}(x_{1},x_{2})$ can be expressed as

\begin{equation}\label{eq:joint-cdf-integral}
F_{X_{1},X_{2}}(x_{1},x_{2}) = \int_{-\infty}^{x_{1}} \int_{-\infty}^{x_{2}} f_{X_{1},X_{2}}(w_{1},w_{2}) \, dw_{1} \, dw_{2},
\end{equation}

for all $(x_{1},x_{2}) \in R^{2}$ . We call the integrand the joint probability density function (pdf) of $(X_{1},X_{2})$ . Then

\[
\frac{\partial^{2} F_{X_{1},X_{2}}(x_{1},x_{2})}{\partial x_{1} \partial x_{2}} = f_{X_{1},X_{2}}(x_{1},x_{2}),
\]

except possibly on events that have probability zero. A pdf is essentially characterized by the two properties

\begin{equation}\label{eq:pdf-properties}
\text{(i) } f_{X_{1},X_{2}}(x_{1},x_{2}) \geq 0 \text{ and (ii)} \int\int_{\mathcal{D}} f_{X_{1},X_{2}}(x_{1},x_{2}) \, dx_{1} \, dx_{2} = 1.
\end{equation}

For an event $A \in \mathcal{D}$ , we have

\[
P[(X_{1},X_{2}) \in A] = \int\int_{A} f_{X_{1},X_{2}}(x_{1},x_{2}) \, dx_{1} \, dx_{2}.
\]

Note that $P[(X_{1},X_{2}) \in A]$ is just the volume under the surface $z = f_{X_{1},X_{2}}(x_{1},x_{2})$ over the set $A$ .

\begin{remark}\label{rem:notation-abbreviations}
As with univariate random variables, we often drop the subscript $(X_{1},X_{2})$ from joint cdfs, pdfs, and pmfs, when it is clear from the context. We also use notation such as $f_{12}$ instead of $f_{X_{1},X_{2}}$ . Besides $(X_{1},X_{2})$ , we often use $(X,Y)$ to express random vectors.
\end{remark}

\begin{example}\label{ex:uniform-circle}
Consider a continuous random vector $(X,Y)$ which is uniformly distributed over the unit circle in $R^{2}$ . Since the area of the unit circle is $\pi$ , the joint pdf is

\[
f(x,y) = \begin{cases}
\frac{1}{\pi} & -1 < y < 1, -\sqrt{1-y^{2}} < x < \sqrt{1-y^{2}} \\
0 & \text{elsewhere}.
\end{cases}
\]

Probabilities of certain events follow immediately from geometry. For instance, let $A$ be the interior of the circle with radius $1/2$ . Then $P[(X,Y) \in A] = \pi(1/2)^{2}/\pi = 1/4$ . Next, let $B$ be the ring formed by the concentric circles with the respective radii of $1/2$ and $\sqrt{2}/2$ . Then $P[(X,Y) \in B] = \pi[(\sqrt{2}/2)^{2} - (1/2)^{2}]/\pi = 1/4$ .
\end{example}

\begin{example}\label{ex:battery-lifetimes}
Suppose an electrical component has two batteries. Let $X$ and $Y$ denote the lifetimes in standard units of the respective batteries. Assume that the pdf of $(X,Y)$ is

\[
f(x,y) = \begin{cases}
4xye^{-(x^{2}+y^{2})} & x > 0, y > 0 \\
0 & \text{elsewhere}.
\end{cases}
\]

The surface $z = f(x,y)$ peaks at $(x,y) = (\sqrt{2}/2,\sqrt{2}/2)$ . For instance, the probability that both batteries survive beyond $\sqrt{2}/2$ units is given by

\[
P\!\left(X > \frac{\sqrt{2}}{2}, Y > \frac{\sqrt{2}}{2}\right) = \int_{\sqrt{2}/2}^{\infty} \int_{\sqrt{2}/2}^{\infty} 4xye^{-(x^{2}+y^{2})} \, dx \, dy = (e^{-1/2})^{2} \approx 0.3679.
\]
\end{example}

For a continuous random vector $(X_{1},X_{2})$ , the support of $(X_{1},X_{2})$ contains all points $(x_{1},x_{2})$ for which $f(x_{1},x_{2}) > 0$ . We denote the support of a random vector by $\mathcal{S}$ . As in the univariate case, $\mathcal{S} \subset \mathcal{D}$ .

\subsection{Marginal Distributions}\label{sub:marginal-distributions}

Let $(X_{1},X_{2})$ be a random vector. Then both $X_{1}$ and $X_{2}$ are random variables. We can obtain their distributions in terms of the joint distribution of $(X_{1},X_{2})$ as follows. Recall that the event which defined the cdf of $X_{1}$ at $x_{1}$ is $\{X_{1} \leq x_{1}\}$ . However,

\[
\{X_{1} \leq x_{1}\} = \{X_{1} \leq x_{1}\} \cap \{-\infty < X_{2} < \infty\} = \{X_{1} \leq x_{1}, -\infty < X_{2} < \infty\}.
\]

Taking probabilities, we have

\begin{equation}\label{eq:marginal-cdf}
F_{X_{1}}(x_{1}) = P[X_{1} \leq x_{1}, -\infty < X_{2} < \infty].
\end{equation}

By Theorem 1.3.6 we can write this equation as $F_{X_{1}}(x_{1}) = \lim_{x_{2} \uparrow \infty} F(x_{1},x_{2})$ . Thus we have a relationship between the cdfs, which we can extend to either the pmf or pdf depending on whether $(X_{1},X_{2})$ is discrete or continuous.

First consider the discrete case. Let $\mathcal{D}_{X_{1}}$ be the support of $X_{1}$ . For $x_{1} \in \mathcal{D}_{X_{1}}$ , Equation \eqref{eq:marginal-cdf} is equivalent to

\[
F_{X_{1}}(x_{1}) = \sum_{w_{1} \leq x_{1}} \left\{ \sum_{x_{2}} p_{X_{1},X_{2}}(w_{1},x_{2}) \right\}.
\]

By the uniqueness of cdfs, the quantity in braces must be the pmf of $X_{1}$ evaluated at $w_{1}$ ; that is,

\begin{equation}\label{eq:marginal-pmf}
p_{X_{1}}(x_{1}) = \sum_{x_{2}} p_{X_{1},X_{2}}(x_{1},x_{2}),
\end{equation}

for all $x_{1} \in \mathcal{D}_{X_{1}}$ . Hence, to find the probability that $X_{1}$ is $x_{1}$ , keep $x_{1}$ fixed and sum $p_{X_{1},X_{2}}$ over all of $x_{2}$ . In terms of a tabled joint pmf with rows comprised of $X_{1}$ support values and columns comprised of $X_{2}$ support values, this says that the distribution of $X_{1}$ can be obtained by the marginal sums of the rows. Likewise, the pmf of $X_{2}$ can be obtained by marginal sums of the columns.

\begin{example}\label{ex:chip-drawing}
Consider a random experiment that consists of drawing at random one chip from a bowl containing 10 chips of the same shape and size. Each chip has an ordered pair of numbers on it: one with $(1,1)$ , one with $(2,1)$ , two with $(3,1)$ , one with $(1,2)$ , two with $(2,2)$ , and three with $(3,2)$ . Let the random variables $X_{1}$ and $X_{2}$ be defined as the respective first and second values of the ordered pair. The joint pmf $p(x_{1},x_{2})$ of $X_{1}$ and $X_{2}$ can be given by the following table:

\begin{center}
\begin{tabular}{c|cc|c}
 & $x_{2}=1$ & $x_{2}=2$ & $p_{X_{1}}(x_{1})$ \\ \hline
$x_{1}=1$ & $1/10$ & $1/10$ & $2/10$ \\
$x_{1}=2$ & $1/10$ & $2/10$ & $3/10$ \\
$x_{1}=3$ & $2/10$ & $3/10$ & $5/10$ \\ \hline
$p_{X_{2}}(x_{2})$ & $4/10$ & $6/10$ &
\end{tabular}
\end{center}
\end{example}

We next consider the continuous case. Let $\mathcal{D}_{X_{1}}$ be the support of $X_{1}$ . For $x_{1} \in \mathcal{D}_{X_{1}}$ , Equation \eqref{eq:marginal-cdf} is equivalent to

\[
F_{X_{1}}(x_{1}) = \int_{-\infty}^{x_{1}} \left\{ \int_{-\infty}^{\infty} f_{X_{1},X_{2}}(w_{1},x_{2}) \, dx_{2} \right\} dw_{1}.
\]

By the uniqueness of cdfs, the quantity in braces must be the pdf of $X_{1}$ evaluated at $w_{1}$ ; that is,

\begin{equation}\label{eq:marginal-pdf}
f_{X_{1}}(x_{1}) = \int_{-\infty}^{\infty} f_{X_{1},X_{2}}(x_{1},x_{2}) \, dx_{2},
\end{equation}

for all $x_{1} \in \mathcal{D}_{X_{1}}$ . Hence, in the continuous case the marginal pdf of $X_{1}$ is found by integrating out $x_{2}$ . Similarly, the marginal pdf of $X_{2}$ is found by integrating out $x_{1}$ .

\begin{example}[Example \ref{ex:uniform-circle} continued]\label{ex:uniform-circle-marginal}
Consider the vector of continuous random variables $(X,Y)$ discussed in Example \ref{ex:uniform-circle}. To find the marginal distribution of $X$ , fix $x$ between $-1$ and $1$ and then integrate out $y$ from $-\sqrt{1-x^{2}}$ to $\sqrt{1-x^{2}}$ . Hence, the marginal pdf of $X$ is

\[
f_{X}(x) = \int_{-\sqrt{1-x^{2}}}^{\sqrt{1-x^{2}}} \frac{1}{\pi} \, dy = \frac{2}{\pi}\sqrt{1-x^{2}}, \quad -1 < x < 1.
\]

Although $(X,Y)$ has a joint uniform distribution, the distribution of $X$ is unimodal with peak at 0. Because the joint pdf is symmetric in $x$ and $y$ , the marginal pdf of $Y$ is the same as that of $X$ .
\end{example}

\begin{example}\label{ex:triangular-region}
Let $X_{1}$ and $X_{2}$ have the joint pdf

\[
f(x_{1},x_{2}) = \begin{cases}
x_{1}+x_{2} & 0 < x_{1} < 1, 0 < x_{2} < 1 \\
0 & \text{elsewhere}.
\end{cases}
\]

Notice the space of the random vector is the interior of the square with vertices $(0,0),(1,0),(1,1)$ and $(0,1)$ . The marginal pdf of $X_{1}$ is

\[
f_{1}(x_{1}) = \int_{0}^{1} (x_{1}+x_{2}) \, dx_{2} = x_{1} + \frac{1}{2}, \quad 0 < x_{1} < 1,
\]

zero elsewhere, and the marginal pdf of $X_{2}$ is

\[
f_{2}(x_{2}) = \int_{0}^{1} (x_{1}+x_{2}) \, dx_{1} = \frac{1}{2} + x_{2}, \quad 0 < x_{2} < 1,
\]

zero elsewhere.

Suppose we want to find the probability $P(X_{1}+X_{2} \leq 1)$ . The region of integration is the interior of the triangle with vertices $(0,0)$ , $(1,0)$ and $(0,1)$ . Fixing $x_{1}$ and integrating with respect to $x_{2}$ , we have

\[
P(X_{1}+X_{2} \leq 1) = \int_{0}^{1} \left[ \int_{0}^{1-x_{1}} (x_{1}+x_{2}) \, dx_{2} \right] dx_{1} = \frac{1}{3}.
\]
\end{example}

\begin{example}[Example \ref{ex:battery-lifetimes} continued]\label{ex:battery-marginal}
Recall that the random variables $X$ and $Y$ of Example \ref{ex:battery-lifetimes} were the lifetimes of two batteries. Using the change-of-variable $w = y^{2}$ , the marginal pdf of $X$ is

\[
f_{X}(x) = \int_{0}^{\infty} 4xye^{-(x^{2}+y^{2})} \, dy = 2xe^{-x^{2}} \int_{0}^{\infty} e^{-w} \, dw = 2xe^{-x^{2}},
\]

for $x > 0$ . By the symmetry of $x$ and $y$ in the model, the pdf of $Y$ is the same as that of $X$ . To determine the median lifetime, $\theta$ , of these batteries, we need to solve

\[
\frac{1}{2} = \int_{0}^{\theta} 2xe^{-x^{2}} \, dx = 1 - e^{-\theta^{2}}.
\]

Solving this equation, we obtain $\theta = \sqrt{\log 2} \approx 0.8326$ .
\end{example}

\subsection{Expectation}\label{sub:expectation-two-rv}

The concept of expectation extends in a straightforward manner. Let $(X_{1},X_{2})$ be a random vector and let $Y = g(X_{1},X_{2})$ for some real-valued function; i.e., $g:R^{2} \to R$ . Then $Y$ is a random variable and we could determine its expectation by obtaining the distribution of $Y$ . But Theorem 1.8.1 is true for random vectors also.

Suppose $(X_{1},X_{2})$ is of the continuous type. Then $E(Y)$ exists if

\[
\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} |g(x_{1},x_{2})| f_{X_{1},X_{2}}(x_{1},x_{2}) \, dx_{1} \, dx_{2} < \infty .
\]

Then

\begin{equation}\label{eq:expectation-continuous}
E(Y) = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} g(x_{1},x_{2}) f_{X_{1},X_{2}}(x_{1},x_{2}) \, dx_{1} \, dx_{2}.
\end{equation}

Likewise if $(X_{1},X_{2})$ is discrete, then $E(Y)$ exists if

\[
\sum_{x_{1}} \sum_{x_{2}} |g(x_{1},x_{2})| p_{X_{1},X_{2}}(x_{1},x_{2}) < \infty .
\]

Then

\begin{equation}\label{eq:expectation-discrete}
E(Y) = \sum_{x_{1}} \sum_{x_{2}} g(x_{1},x_{2}) p_{X_{1},X_{2}}(x_{1},x_{2}).
\end{equation}

We can now show that $E$ is a linear operator.

\begin{theorem}\label{thm:linearity-expectation}
Let $(X_{1},X_{2})$ be a random vector. Let $Y_{1} = g_{1}(X_{1},X_{2})$ and $Y_{2} = g_{2}(X_{1},X_{2})$ be random variables whose expectations exist. Then for all real numbers $k_{1}$ and $k_{2}$ ,

\begin{equation}\label{eq:linearity-expectation}
E(k_{1}Y_{1}+k_{2}Y_{2}) = k_{1}E(Y_{1}) + k_{2}E(Y_{2}).
\end{equation}
\end{theorem}

\begin{example}\label{ex:expectation-computation}
Let $X_{1}$ and $X_{2}$ have the pdf

\[
f(x_{1},x_{2}) = \begin{cases}
8x_{1}x_{2} & 0 < x_{1} < x_{2} < 1 \\
0 & \text{elsewhere}.
\end{cases}
\]

Then

\[
E(X_{1}X_{2}^{2}) = \int_{0}^{1} \left[ \int_{0}^{x_{2}} 8x_{1}^{2}x_{2}^{3} \, dx_{1} \right] dx_{2} = \int_{0}^{1} \frac{8}{3}x_{2}^{6} \, dx_{2} = \frac{8}{21}.
\]

In addition,

\[
E(X_{2}) = \int_{0}^{1} \left[ \int_{0}^{x_{2}} x_{2}(8x_{1}x_{2}) \, dx_{1} \right] dx_{2} = \frac{4}{5}.
\]

Since $X_{2}$ has the pdf $f_{2}(x_{2}) = 4x_{2}^{3}$ , $0 < x_{2} < 1$ , zero elsewhere, the latter expectation can also be found by

\[
E(X_{2}) = \int_{0}^{1} x_{2}(4x_{2}^{3}) \, dx_{2} = \frac{4}{5}.
\]

Using Theorem \ref{thm:linearity-expectation},

\[
E(7X_{1}X_{2}^{2} + 5X_{2}) = 7E(X_{1}X_{2}^{2}) + 5E(X_{2}) = 7 \cdot \frac{8}{21} + 5 \cdot \frac{4}{5} = \frac{20}{3}.
\]
\end{example}

We also note that the expected value of any function $g(X_{2})$ of $X_{2}$ can be found in two ways:

\[
E(g(X_{2})) = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} g(x_{2}) f(x_{1},x_{2}) \, dx_{1} \, dx_{2} = \int_{-\infty}^{\infty} g(x_{2}) f_{X_{2}}(x_{2}) \, dx_{2},
\]

the latter single integral being obtained from the double integral by integrating on $x_{1}$ first.

\begin{definition}[Moment Generating Function of a Random Vector]\label{def:mgf-random-vector}
Let $\mathbf{X} = (X_{1},X_{2})'$ be a random vector. If $E(e^{t_{1}X_{1}+t_{2}X_{2}})$ exists for $|t_{1}| < h_{1}$ and $|t_{2}| < h_{2}$ , where $h_{1}$ and $h_{2}$ are positive, it is denoted by $M_{X_{1},X_{2}}(t_{1},t_{2})$ and is called the moment generating function (mgf) of $\mathbf{X}$ .
\end{definition}

As in the one-variable case, if it exists, the mgf of a random vector uniquely determines the distribution of the random vector.

Let $\mathbf{t} = (t_{1},t_{2})'$ . Then we can write the mgf of $\mathbf{X}$ as

\begin{equation}\label{eq:mgf-vector}
M_{X_{1},X_{2}}(\mathbf{t}) = E\!\left[ e^{\mathbf{t}'\mathbf{X}} \right].
\end{equation}

Also, the mgfs of $X_{1}$ and $X_{2}$ are immediately seen to be $M_{X_{1},X_{2}}(t_{1},0)$ and $M_{X_{1},X_{2}}(0,t_{2})$ , respectively.

\begin{example}\label{ex:mgf-computation}
Let the continuous-type random variables $X$ and $Y$ have the joint pdf

\[
f(x,y) = \begin{cases}
e^{-y} & 0 < x < y < \infty \\
0 & \text{elsewhere}.
\end{cases}
\]

The mgf of this joint distribution is

\[
M(t_{1},t_{2}) = \int_{0}^{\infty} \left[ \int_{x}^{\infty} \exp(t_{1}x + t_{2}y - y) \, dy \right] dx = \frac{1}{(1-t_{1}-t_{2})(1-t_{2})},
\]

provided that $t_{1}+t_{2} < 1$ and $t_{2} < 1$ . Furthermore, the moment-generating functions of the marginal distributions of $X$ and $Y$ are, respectively,

\[
M(t_{1},0) = \frac{1}{1-t_{1}}, \quad t_{1} < 1,
\]
\[
M(0,t_{2}) = \frac{1}{(1-t_{2})^{2}}, \quad t_{2} < 1.
\]

These are respectively those of the marginal probability density functions

\[
f_{1}(x) = \int_{x}^{\infty} e^{-y} \, dy = e^{-x}, \quad 0 < x < \infty ,
\]

zero elsewhere, and

\[
f_{2}(y) = e^{-y}\int_{0}^{y} dx = ye^{-y}, \quad 0 < y < \infty ,
\]

zero elsewhere.
\end{example}

\begin{definition}[Expected Value of a Random Vector]\label{def:expected-value-vector}
Let $\mathbf{X} = (X_{1},X_{2})'$ be a random vector. Then the expected value of $\mathbf{X}$ exists if the expectations of $X_{1}$ and $X_{2}$ exist. If it exists, then the expected value is given by

\begin{equation}\label{eq:expected-value-vector}
E[\mathbf{X}] = \begin{bmatrix} E(X_{1}) \\ E(X_{2}) \end{bmatrix}.
\end{equation}
\end{definition}

\section{Transformations}\label{sec:transformations-two-rv}

Let $(X_{1},X_{2})$ be a random vector with known distribution. Often we are interested in a function of $(X_{1},X_{2})$ such as $Y = u(X_{1},X_{2})$ . In this section we discuss the transformation technique for finding the distribution of $Y$ . We begin with a discussion of one-to-one transformations of random vectors.

Suppose $(X_{1},X_{2})$ is a continuous type random vector with pdf $f_{X_{1},X_{2}}(x_{1},x_{2})$ defined on the support $\mathcal{S}$ . Let $Y_{1} = u_{1}(X_{1},X_{2})$ and $Y_{2} = u_{2}(X_{1},X_{2})$ define a transformation that maps $\mathcal{S}$ onto $\mathcal{T}$ . If this transformation is one-to-one, then there exists an inverse transformation $X_{1} = v_{1}(Y_{1},Y_{2})$ and $X_{2} = v_{2}(Y_{1},Y_{2})$ that maps $\mathcal{T}$ onto $\mathcal{S}$ . The joint pdf of $(Y_{1},Y_{2})$ is given by

\begin{equation}\label{eq:transformation-joint-pdf}
f_{Y_{1},Y_{2}}(y_{1},y_{2}) = f_{X_{1},X_{2}}(v_{1}(y_{1},y_{2}), v_{2}(y_{1},y_{2})) \cdot |J|,
\end{equation}

where $J$ is the Jacobian of the transformation:

\[
J = \det\begin{pmatrix}
\frac{\partial x_{1}}{\partial y_{1}} & \frac{\partial x_{1}}{\partial y_{2}} \\
\frac{\partial x_{2}}{\partial y_{1}} & \frac{\partial x_{2}}{\partial y_{2}}
\end{pmatrix}
= \frac{\partial x_{1}}{\partial y_{1}} \cdot \frac{\partial x_{2}}{\partial y_{2}} - \frac{\partial x_{1}}{\partial y_{2}} \cdot \frac{\partial x_{2}}{\partial y_{1}}.
\]

\begin{example}\label{ex:transformation-sum-difference}
Let $X_{1}$ and $X_{2}$ have the joint pdf

\[
f_{X_{1},X_{2}}(x_{1},x_{2}) = \begin{cases}
\frac{1}{8} & -2 < x_{1} < 2, -1 < x_{2} < 1 \\
0 & \text{elsewhere}.
\end{cases}
\]

Suppose we want to find the joint pdf of $Y_{1} = X_{1}+X_{2}$ and $Y_{2} = X_{1}-X_{2}$ . The inverse transformation is $x_{1} = (y_{1}+y_{2})/2$ and $x_{2} = (y_{1}-y_{2})/2$ . The Jacobian is

\[
J = \det\begin{pmatrix}
1/2 & 1/2 \\
1/2 & -1/2
\end{pmatrix} = -\frac{1}{2} - \frac{1}{2} = -1.
\]

The support $\mathcal{T}$ is determined by $|y_{1}+y_{2}| < 4$ and $|y_{1}-y_{2}| < 2$ . Hence, the joint pdf of $(Y_{1},Y_{2})$ is

\[
f_{Y_{1},Y_{2}}(y_{1},y_{2}) = \frac{1}{8} \cdot |J| = \frac{1}{8},
\]

over the region $\mathcal{T}$ .
\end{example}

\begin{example}\label{ex:transformation-product-quotient}
Let $X_{1}$ and $X_{2}$ have the joint pdf

\[
f_{X_{1},X_{2}}(x_{1},x_{2}) = \begin{cases}
8x_{1}x_{2} & 0 < x_{1} < x_{2} < 1 \\
0 & \text{elsewhere}.
\end{cases}
\]

Suppose $Y_{1} = X_{1}/X_{2}$ and $Y_{2} = X_{2}$ . The inverse transformation is $x_{1} = y_{1}y_{2}$ and $x_{2} = y_{2}$ . The Jacobian is $J = y_{2}$ . The support of $(Y_{1},Y_{2})$ is $\mathcal{T} = \{(y_{1},y_{2}) : 0 < y_{1} < 1, 0 < y_{2} < 1\}$ . Hence, the joint pdf is

\[
f_{Y_{1},Y_{2}}(y_{1},y_{2}) = 8y_{1}y_{2} \cdot y_{2} = 8y_{1}y_{2}^{3},
\]

over $\mathcal{T}$ . The marginal pdfs are

\[
f_{Y_{1}}(y_{1}) = \int_{0}^{1} 8y_{1}y_{2}^{3} \, dy_{2} = 2y_{1}, \quad 0 < y_{1} < 1,
\]

and

\[
f_{Y_{2}}(y_{2}) = \int_{0}^{1} 8y_{1}y_{2}^{3} \, dy_{1} = 4y_{2}^{3}, \quad 0 < y_{2} < 1.
\]
\end{example}

In addition to the change-of-variable and cdf techniques for finding distributions of functions of random variables, there is another method, called the moment generating function (mgf) technique, which works well for linear functions of random variables.

\begin{example}\label{ex:mgf-technique-sum}
Let $X_{1}$ and $X_{2}$ have the joint pmf

\[
p_{X_{1},X_{2}}(x_{1},x_{2}) = \begin{cases}
\frac{\mu_{1}^{x_{1}}\mu_{2}^{x_{2}} e^{-\mu_{1}} e^{-\mu_{2}}}{x_{1}! \, x_{2}!} & x_{1}=0,1,2,\ldots, x_{2}=0,1,2,\ldots \\
0 & \text{elsewhere}.
\end{cases}
\]

Let $Y = X_{1}+X_{2}$ and consider

\[
E(e^{tY}) = \sum_{x_{1}=0}^{\infty} \sum_{x_{2}=0}^{\infty} e^{t(x_{1}+x_{2})} p_{X_{1},X_{2}}(x_{1},x_{2}) = e^{(\mu_{1}+\mu_{2})(e^{t}-1)}.
\]

By the uniqueness of mgfs, the pmf of $Y$ must be

\[
p_{Y}(y) = e^{-(\mu_{1}+\mu_{2})} \frac{(\mu_{1}+\mu_{2})^{y}}{y!}, \quad y=0,1,2,\ldots,
\]

which is a Poisson distribution with parameter $\mu_{1}+\mu_{2}$ .
\end{example}

\section{Conditional Distributions and Expectations}\label{sec:conditional-distributions}

In Section \ref{sec:distributions-two-rv} we introduced the joint probability distribution of a pair of random variables. We also showed how to recover the individual (marginal) distributions for the random variables from the joint distribution. In this section, we discuss conditional distributions, i.e., the distribution of one of the random variables when the other has assumed a specific value.

Let $X_{1}$ and $X_{2}$ denote random variables of the discrete type, which have the joint pmf $p_{X_{1},X_{2}}(x_{1},x_{2})$ that is positive on the support set $\mathcal{S}$ and is zero elsewhere. Let $p_{X_{1}}(x_{1})$ and $p_{X_{2}}(x_{2})$ denote, respectively, the marginal probability mass functions of $X_{1}$ and $X_{2}$ . Let $x_{1}$ be a point in the support of $X_{1}$ ; hence, $p_{X_{1}}(x_{1}) > 0$ . Using the definition of conditional probability, we have

\[
P(X_{2}=x_{2} \mid X_{1}=x_{1}) = \frac{P(X_{1}=x_{1},X_{2}=x_{2})}{P(X_{1}=x_{1})} = \frac{p_{X_{1},X_{2}}(x_{1},x_{2})}{p_{X_{1}}(x_{1})}.
\]

Define this function as

\begin{equation}\label{eq:conditional-pmf}
p_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) = \frac{p_{X_{1},X_{2}}(x_{1},x_{2})}{p_{X_{1}}(x_{1})}, \quad x_{2} \in \mathcal{S}_{X_{2}}.
\end{equation}

For any fixed $x_{1}$ with $p_{X_{1}}(x_{1}) > 0$ , this function $p_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ satisfies the conditions of being a pmf of the discrete type because $p_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ is nonnegative and

\[
\sum_{x_{2}} p_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) = \frac{1}{p_{X_{1}}(x_{1})} \sum_{x_{2}} p_{X_{1},X_{2}}(x_{1},x_{2}) = \frac{p_{X_{1}}(x_{1})}{p_{X_{1}}(x_{1})} = 1.
\]

We call $p_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ the conditional pmf of the discrete type of random variable $X_{2}$ , given that the discrete type of random variable $X_{1} = x_{1}$ .

Now let $X_{1}$ and $X_{2}$ denote random variables of the continuous type and have the joint pdf $f_{X_{1},X_{2}}(x_{1},x_{2})$ and the marginal probability density functions $f_{X_{1}}(x_{1})$ and $f_{X_{2}}(x_{2})$ , respectively. When $f_{X_{1}}(x_{1}) > 0$ , we define the symbol $f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ by the relation

\begin{equation}\label{eq:conditional-pdf}
f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) = \frac{f_{X_{1},X_{2}}(x_{1},x_{2})}{f_{X_{1}}(x_{1})}.
\end{equation}

In this relation, $x_{1}$ is to be thought of as having a fixed (but any fixed) value for which $f_{X_{1}}(x_{1}) > 0$ . It is evident that $f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ is nonnegative and that

\[
\int_{-\infty}^{\infty} f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) \, dx_{2} = \frac{1}{f_{X_{1}}(x_{1})} \int_{-\infty}^{\infty} f_{X_{1},X_{2}}(x_{1},x_{2}) \, dx_{2} = \frac{f_{X_{1}}(x_{1})}{f_{X_{1}}(x_{1})} = 1.
\]

That is, $f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ has the properties of a pdf of one continuous type of random variable. It is called the conditional pdf of the continuous type of random variable $X_{2}$ , given that the continuous type of random variable $X_{1}$ has the value $x_{1}$ .

Since each of $f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ and $f_{X_{1}\mid X_{2}}(x_{1}\mid x_{2})$ is a pdf of one random variable, each has all the properties of such a pdf. If $u(X_{2})$ is a function of $X_{2}$ , the conditional expectation of $u(X_{2})$ , given that $X_{1} = x_{1}$ , if it exists, is given by

\begin{equation}\label{eq:conditional-expectation}
E[u(X_{2}) \mid x_{1}] = \int_{-\infty}^{\infty} u(x_{2}) f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) \, dx_{2}.
\end{equation}

Note that $E[u(X_{2})\mid x_{1}]$ is a function of $x_{1}$ . If they do exist, then $E(X_{2}\mid x_{1})$ is the mean and $\operatorname{Var}(X_{2}\mid x_{1})$ is the conditional variance of the conditional distribution of $X_{2}$ , given $X_{1} = x_{1}$ . Of course, we have

\[
\operatorname{Var}(X_{2}\mid x_{1}) = E(X_{2}^{2}\mid x_{1}) - [E(X_{2}\mid x_{1})]^{2}.
\]

\begin{example}\label{ex:conditional-distribution-triangular}
Let $X_{1}$ and $X_{2}$ have the joint pdf

\[
f(x_{1},x_{2}) = \begin{cases}
2 & 0 < x_{1} < x_{2} < 1 \\
0 & \text{elsewhere}.
\end{cases}
\]

Then the marginal probability density functions are, respectively,

\[
f_{1}(x_{1}) = \begin{cases}
\int_{x_{1}}^{1} 2 \, dx_{2} = 2(1-x_{1}) & 0 < x_{1} < 1 \\
0 & \text{elsewhere},
\end{cases}
\]

and

\[
f_{2}(x_{2}) = \begin{cases}
\int_{0}^{x_{2}} 2 \, dx_{1} = 2x_{2} & 0 < x_{2} < 1 \\
0 & \text{elsewhere}.
\end{cases}
\]

The conditional pdf of $X_{1}$ , given $X_{2} = x_{2}$ , $0 < x_{2} < 1$ , is

\[
f_{1\mid 2}(x_{1}\mid x_{2}) = \frac{2}{2x_{2}} = \frac{1}{x_{2}}, \quad 0 < x_{1} < x_{2} < 1.
\]

Here the conditional mean and the conditional variance of $X_{1}$ , given $X_{2} = x_{2}$ , are respectively,

\[
E(X_{1}\mid x_{2}) = \int_{0}^{x_{2}} x_{1} \cdot \frac{1}{x_{2}} \, dx_{1} = \frac{x_{2}}{2}, \quad 0 < x_{2} < 1,
\]

and

\[
\operatorname{Var}(X_{1}\mid x_{2}) = \int_{0}^{x_{2}} \left(x_{1}-\frac{x_{2}}{2}\right)^{2} \cdot \frac{1}{x_{2}} \, dx_{1} = \frac{x_{2}^{2}}{12}, \quad 0 < x_{2} < 1.
\]
\end{example}

\begin{theorem}\label{ex:law-total-expectation}
Let $(X_{1},X_{2})$ be a random vector such that the variance of $X_{2}$ is finite. Then:

\begin{enumerate}
\item $E[E(X_{2}\mid X_{1})] = E(X_{2})$ .
\item $\operatorname{Var}[E(X_{2}\mid X_{1})] \leq \operatorname{Var}(X_{2})$ .
\end{enumerate}
\end{theorem}

\begin{remark}\label{rem:rao-blackwell}
Intuitively, this result could have this useful interpretation. Both the random variables $X_{2}$ and $E(X_{2}\mid X_{1})$ have the same mean $\mu_{2}$ . If we did not know $\mu_{2}$ , we could use either of the two random variables to guess at the unknown $\mu_{2}$ . Since, however, $\operatorname{Var}(X_{2}) \geq \operatorname{Var}[E(X_{2}\mid X_{1})]$ , we would put more reliance on $E(X_{2}\mid X_{1})$ as a guess. That is, if we observe the pair $(X_{1},X_{2})$ to be $(x_{1},x_{2})$ , we could prefer to use $E(X_{2}\mid x_{1})$ to $x_{2}$ as a guess at the unknown $\mu_{2}$ . When studying the use of sufficient statistics in estimation in Chapter 7, we make use of this famous result, attributed to C. R. Rao and David Blackwell.
\end{remark}

\section{Independent Random Variables}\label{sec:independent-random-variables}

Let $X_{1}$ and $X_{2}$ denote the random variables of the continuous type that have the joint pdf $f(x_{1},x_{2})$ and marginal probability density functions $f_{1}(x_{1})$ and $f_{2}(x_{2})$ , respectively. In accordance with the definition of the conditional pdf $f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ , we may write the joint pdf $f(x_{1},x_{2})$ as

\[
f(x_{1},x_{2}) = f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) f_{1}(x_{1}).
\]

Suppose that we have an instance where $f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ does not depend upon $x_{1}$ . Then the marginal pdf of $X_{2}$ is, for random variables of the continuous type,

\[
f_{2}(x_{2}) = \int_{-\infty}^{\infty} f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) f_{1}(x_{1}) \, dx_{1} = f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) \int_{-\infty}^{\infty} f_{1}(x_{1}) \, dx_{1} = f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}).
\]

Accordingly,

\[
f_{2}(x_{2}) = f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) \quad \text{and} \quad f(x_{1},x_{2}) = f_{1}(x_{1})f_{2}(x_{2}),
\]

when $f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1})$ does not depend upon $x_{1}$ . That is, if the conditional distribution of $X_{2}$ , given $X_{1} = x_{1}$ , is independent of any assumption about $x_{1}$ , then $f(x_{1},x_{2}) = f_{1}(x_{1})f_{2}(x_{2})$ .

\begin{definition}[Independence]\label{def:independence}
Let the random variables $X_{1}$ and $X_{2}$ have the joint pdf $f(x_{1},x_{2})$ [joint pmf $p(x_{1},x_{2})$ ] and the marginal pdfs [pmfs] $f_{1}(x_{1})$ $[p_{1}(x_{1})]$ and $f_{2}(x_{2})$ $[p_{2}(x_{2})]$ , respectively. The random variables $X_{1}$ and $X_{2}$ are said to be independent if, and only if, $f(x_{1},x_{2}) \equiv f_{1}(x_{1})f_{2}(x_{2})$ $[p(x_{1},x_{2}) \equiv p_{1}(x_{1})p_{2}(x_{2})]$ . Random variables that are not independent are said to be dependent.
\end{definition}

\begin{remark}\label{rem:independence-product-space}
Two comments should be made about the preceding definition. First, the product of two positive functions $f_{1}(x_{1})f_{2}(x_{2})$ means a function that is positive on the product space. That is, if $f_{1}(x_{1})$ and $f_{2}(x_{2})$ are positive on, and only on, the respective spaces $\mathcal{S}_{1}$ and $\mathcal{S}_{2}$ , then the product of $f_{1}(x_{1})$ and $f_{2}(x_{2})$ is positive on, and only on, the product space $\mathcal{S} = \{(x_{1},x_{2}) : x_{1} \in \mathcal{S}_{1}, x_{2} \in \mathcal{S}_{2}\}$ .
\end{remark}

\begin{example}\label{ex:dependent-demonstration}
Let the joint pdf of $X_{1}$ and $X_{2}$ be

\[
f(x_{1},x_{2}) = \begin{cases}
x_{1}+x_{2} & 0 < x_{1} < 1, 0 < x_{2} < 1 \\
0 & \text{elsewhere}.
\end{cases}
\]

We show that $X_{1}$ and $X_{2}$ are dependent. Here the marginal probability density functions are

\[
f_{1}(x_{1}) = \begin{cases}
\int_{0}^{1} (x_{1}+x_{2}) \, dx_{2} = x_{1}+\frac{1}{2} & 0 < x_{1} < 1 \\
0 & \text{elsewhere},
\end{cases}
\]

and

\[
f_{2}(x_{2}) = \begin{cases}
\int_{0}^{1} (x_{1}+x_{2}) \, dx_{1} = \frac{1}{2}+x_{2} & 0 < x_{2} < 1 \\
0 & \text{elsewhere}.
\end{cases}
\]

Since $f(x_{1},x_{2}) \neq f_{1}(x_{1})f_{2}(x_{2})$ , the random variables $X_{1}$ and $X_{2}$ are dependent.
\end{example}

\begin{theorem}\label{ex:independence-factorization}
Let the random variables $X_{1}$ and $X_{2}$ have supports $\mathcal{S}_{1}$ and $\mathcal{S}_{2}$ , respectively, and have the joint pdf $f(x_{1},x_{2})$ . Then $X_{1}$ and $X_{2}$ are independent if and only if $f(x_{1},x_{2})$ can be written as a product of a nonnegative function of $x_{1}$ and a nonnegative function of $x_{2}$ . That is,

\[
f(x_{1},x_{2}) \equiv g(x_{1})h(x_{2}),
\]

where $g(x_{1}) > 0$ , $x_{1} \in \mathcal{S}_{1}$ , zero elsewhere, and $h(x_{2}) > 0$ , $x_{2} \in \mathcal{S}_{2}$ , zero elsewhere.
\end{theorem}

The following theorem provides a useful criterion for establishing independence.

\begin{theorem}\label{ex:independence-cdf}
Let the random variables $X_{1}$ and $X_{2}$ have the joint cdf $F(x_{1},x_{2})$ and marginal cdfs $F_{1}(x_{1})$ and $F_{2}(x_{2})$ , respectively. Then $X_{1}$ and $X_{2}$ are independent if and only if

\[
F(x_{1},x_{2}) \equiv F_{1}(x_{1})F_{2}(x_{2}).
\]

In the discrete case, this is equivalent to $p(x_{1},x_{2}) \equiv p_{1}(x_{1})p_{2}(x_{2})$ . In the continuous case, this is equivalent to $f(x_{1},x_{2}) \equiv f_{1}(x_{1})f_{2}(x_{2})$ .
\end{theorem}

If $X_{1}$ and $X_{2}$ are independent, then the conditional distributions equal the marginal distributions:

\[
f_{X_{2}\mid X_{1}}(x_{2}\mid x_{1}) = f_{2}(x_{2}) \quad \text{and} \quad f_{X_{1}\mid X_{2}}(x_{1}\mid x_{2}) = f_{1}(x_{1}).
\]

Moreover, for any functions $g(X_{1})$ and $h(X_{2})$ ,

\[
E[g(X_{1})h(X_{2})] = E[g(X_{1})] \cdot E[h(X_{2})].
\]

In particular, if $X_{1}$ and $X_{2}$ are independent, then

\[
\operatorname{Cov}(X_{1},X_{2}) = 0.
\]

However, the converse is not true in general; zero covariance does not imply independence (unless the distribution is bivariate normal).

\section{Correlation Coefficient}\label{sec:correlation-coefficient}

In Section \ref{sec:distributions-two-rv} we defined the expected value of a random vector. Now we discuss measures of linear association between two random variables.

\begin{definition}[Covariance]\label{def:covariance}
The covariance between $X_{1}$ and $X_{2}$ is defined as

\[
\operatorname{Cov}(X_{1},X_{2}) = E[(X_{1}-E[X_{1}])(X_{2}-E[X_{2}])] = E(X_{1}X_{2}) - E(X_{1})E(X_{2}).
\]
\end{definition}

The covariance has the following properties:

\begin{enumerate}
\item $\operatorname{Cov}(X_{1},X_{1}) = \operatorname{Var}(X_{1})$ .
\item $\operatorname{Cov}(X_{1},X_{2}) = \operatorname{Cov}(X_{2},X_{1})$ .
\item $\operatorname{Cov}(aX_{1},X_{2}) = a\operatorname{Cov}(X_{1},X_{2})$ .
\item $\operatorname{Cov}(X_{1}+X_{2},Y) = \operatorname{Cov}(X_{1},Y) + \operatorname{Cov}(X_{2},Y)$ .
\item If $X_{1}$ and $X_{2}$ are independent, then $\operatorname{Cov}(X_{1},X_{2}) = 0$ .
\end{enumerate}

From property (4), we have

\[
\operatorname{Var}(X_{1}+X_{2}) = \operatorname{Var}(X_{1}) + \operatorname{Var}(X_{2}) + 2\operatorname{Cov}(X_{1},X_{2}).
\]

\begin{definition}[Correlation Coefficient]\label{def:correlation-coefficient}
The correlation coefficient between $X_{1}$ and $X_{2}$ is defined as

\[
\rho_{12} = \text{corr}(X_{1},X_{2}) = \frac{\operatorname{Cov}(X_{1},X_{2})}{\sqrt{\operatorname{Var}(X_{1})} \sqrt{\operatorname{Var}(X_{2})}} = \frac{\operatorname{Cov}(X_{1},X_{2})}{\sigma_{1}\sigma_{2}},
\]

provided $\sigma_{1} > 0$ and $\sigma_{2} > 0$ .
\end{definition}

The correlation coefficient satisfies $-1 \leq \rho_{12} \leq 1$ . Moreover, $|\rho_{12}| = 1$ if and only if $X_{1}$ and $X_{2}$ are linearly related; that is, there exist constants $a \neq 0$ and $b$ such that $X_{2} = aX_{1}+b$ with probability 1. When $\rho_{12} = 0$ , we say that $X_{1}$ and $X_{2}$ are uncorrelated.

\begin{remark}\label{rem:covariance-vs-correlation}
Note that covariance depends on the units of measurement, while the correlation coefficient is dimensionless. The correlation coefficient is a standardized measure of linear association.
\end{remark}

If $X_{1}$ and $X_{2}$ are independent, then they are uncorrelated ($\rho_{12}=0$ ). However, the converse is not generally true; uncorrelated random variables need not be independent.

\section{The Bivariate Normal Distribution}\label{sec:bivariate-normal}

One of the most important joint distributions is the bivariate normal distribution.

\begin{definition}[Bivariate Normal Distribution]\label{def:bivariate-normal}
We say that $(X_{1},X_{2})$ has a bivariate normal distribution if the joint pdf is given by

\[
f(x_{1},x_{2}) = \frac{1}{2\pi\sigma_{1}\sigma_{2}\sqrt{1-\rho^{2}}} \exp\left\{ -\frac{1}{2(1-\rho^{2})} \left[ \frac{(x_{1}-\mu_{1})^{2}}{\sigma_{1}^{2}} - 2\rho\frac{(x_{1}-\mu_{1})(x_{2}-\mu_{2})}{\sigma_{1}\sigma_{2}} + \frac{(x_{2}-\mu_{2})^{2}}{\sigma_{2}^{2}} \right] \right\},
\]

for $-\infty < x_{1} < \infty$ , $-\infty < x_{2} < \infty$ , where $\mu_{1} \in R$ , $\mu_{2} \in R$ , $\sigma_{1} > 0$ , $\sigma_{2} > 0$ , and $-1 < \rho < 1$ . We write $(X_{1},X_{2}) \sim N(\mu_{1},\mu_{2},\sigma_{1}^{2},\sigma_{2}^{2},\rho)$ .
\end{definition}

The parameters have the following interpretations:
\begin{itemize}
\item $E(X_{1}) = \mu_{1}$ and $E(X_{2}) = \mu_{2}$ .
\item $\operatorname{Var}(X_{1}) = \sigma_{1}^{2}$ and $\operatorname{Var}(X_{2}) = \sigma_{2}^{2}$ .
\item $\rho$ is the correlation coefficient between $X_{1}$ and $X_{2}$ .
\end{itemize}

\begin{theorem}[Properties of Bivariate Normal]\label{ex:bivariate-normal-properties}
If $(X_{1},X_{2}) \sim N(\mu_{1},\mu_{2},\sigma_{1}^{2},\sigma_{2}^{2},\rho)$ , then:

\begin{enumerate}
\item The marginal distributions are $X_{1} \sim N(\mu_{1},\sigma_{1}^{2})$ and $X_{2} \sim N(\mu_{2},\sigma_{2}^{2})$ .
\item The conditional distribution of $X_{2}$ given $X_{1}=x_{1}$ is normal with mean and variance:
\[
E(X_{2}\mid x_{1}) = \mu_{2} + \rho\frac{\sigma_{2}}{\sigma_{1}}(x_{1}-\mu_{1}), \quad \operatorname{Var}(X_{2}\mid x_{1}) = \sigma_{2}^{2}(1-\rho^{2}).
\]
\item $X_{1}$ and $X_{2}$ are independent if and only if $\rho = 0$ .
\item Any linear combination $aX_{1}+bX_{2}$ has a normal distribution.
\end{enumerate}
\end{theorem}

A remarkable property of the bivariate normal distribution is that zero covariance (uncorrelatedness) implies independence.

\begin{corollary}\label{cor:bivariate-normal-uncorrelated-independent}
If $(X_{1},X_{2})$ has a bivariate normal distribution and $\rho = 0$ , then $X_{1}$ and $X_{2}$ are independent.
\end{corollary}

\section{Exercises}\label{sec:chapter2-exercises}

\begin{exercise}{2.1.1}
Let $f(x_{1},x_{2}) = 4x_{1}x_{2}$ , $0 < x_{1} < 1$ , $0 < x_{2} < 1$ , zero elsewhere, be the pdf of $X_{1}$ and $X_{2}$ . Find $P(0 < X_{1} < \frac{1}{2},\frac{1}{4} < X_{2} < 1)$ , $P(X_{1} = X_{2})$ , $P(X_{1} < X_{2})$ , and $P(X_{1}\leq X_{2})$ .
\end{exercise}

\begin{exercise}{2.1.3}
Let $F(x,y)$ be the distribution function of $X$ and $Y$ . For all real constants $a < b$ , $c < d$ , show that
\[
P(a < X \leq b, c < Y \leq d) = F(b,d) - F(b,c) - F(a,d) + F(a,c).
\]
\end{exercise}

\begin{exercise}{2.1.7}
Let $f(x,y) = e^{-x-y}$ , $0 < x < \infty$ , $0 < y < \infty$ , zero elsewhere, be the pdf of $X$ and $Y$ . Then if $Z = X+Y$ , compute $P(Z \leq 0)$ , $P(Z \leq 6)$ , and, more generally, $P(Z \leq z)$ , for $0 < z < \infty$ . What is the pdf of $Z$ ?
\end{exercise}

\begin{exercise}{2.1.11}
Let $X_{1}$ and $X_{2}$ have the joint pdf $f(x_{1},x_{2}) = 15x_{1}^{2}x_{2}$ , $0 < x_{1} < x_{2} < 1$ , zero elsewhere. Find the marginal pdfs and compute $P(X_{1}+X_{2}\leq 1)$ .
\end{exercise}

\begin{exercise}{2.2.5}
Let $X_{1}$ and $X_{2}$ be continuous random variables with the joint probability density function $f_{X_{1},X_{2}}(x_{1},x_{2}), -\infty < x_{i} < \infty, i=1,2$ . Let $Y_{1} = X_{1}+X_{2}$ and $Y_{2} = X_{2}$ .

\begin{enumerate}
\item Find the joint pdf $f_{Y_{1},Y_{2}}$ .
\item Show that
\[
f_{Y_{1}}(y_{1}) = \int_{-\infty}^{\infty} f_{X_{1},X_{2}}(y_{1}-y_{2},y_{2}) \, dy_{2},
\]
which is sometimes called the convolution formula.
\end{enumerate}
\end{exercise}

\begin{exercise}{2.3.1}
Let $X_{1}$ and $X_{2}$ have the joint pdf $f(x_{1},x_{2}) = x_{1}+x_{2}$ , $0 < x_{1} < 1$ , $0 < x_{2} < 1$ , zero elsewhere. Find the conditional mean and variance of $X_{2}$ , given $X_{1}=x_{1}$ , $0 < x_{1} < 1$ .
\end{exercise}

\begin{exercise}{2.4.1}
Let the joint pdf of $X_{1}$ and $X_{2}$ be
\[
f(x_{1},x_{2}) = \begin{cases}
2 & 0 < x_{1} < x_{2} < 1 \\
0 & \text{elsewhere}.
\end{cases}
\]
Are $X_{1}$ and $X_{2}$ independent? Why or why not?
\end{exercise}

\userannotation{Chapter 2 focuses on bivariate distributions: joint pmf/pdf, marginal distributions, conditional distributions, independence, transformations, and the correlation coefficient. Key formulas: (2.1.1) joint cdf, (2.1.8)/(2.1.9) marginal pmf/pdf, (2.1.10)/(2.1.11) expectation, (2.3.1)/(2.3.2) conditional pmf/pdf, (2.3.1)/(2.3.2) conditional expectation.}
