\chapter{Some Elementary Statistical Inferences}\label{chap:elementary-statistical-inferences}

\section{Sampling and Statistics}\label{sec:sampling-statistics}

In Chapter 2, we introduced the concepts of samples and statistics. We continue with this development in this chapter while introducing the main tools of inference: confidence intervals and tests of hypotheses.

In a typical statistical problem, we have a random variable $X$ of interest, but its pdf $f(x)$ or pmf $p(x)$ is not known. Our ignorance about $f(x)$ or $p(x)$ can roughly be classified in one of two ways:

\begin{enumerate}
\item $f(x)$ or $p(x)$ is completely unknown.
\item The form of $f(x)$ or $p(x)$ is known down to a parameter $\theta$ , where $\theta$ may be a vector.
\end{enumerate}

For now, we consider the second classification, although some of our discussion pertains to the first classification also. Some examples are the following:

\begin{enumerate}
\item[(a)] $X$ has an exponential distribution $\operatorname{Exp}(\theta)$ , where $\theta$ is unknown.
\item[(b)] $X$ has a binomial distribution $b(n,p)$ , where $n$ is known but $p$ is unknown.
\item[(c)] $X$ has a gamma distribution $\Gamma(\alpha,\beta)$ , where both $\alpha$ and $\beta$ are unknown.
\item[(d)] $X$ has a normal distribution $N(\mu,\sigma^{2})$ , where both the mean $\mu$ and the variance $\sigma^{2}$ of $X$ are unknown.
\end{enumerate}

We often denote this problem by saying that the random variable $X$ has a density or mass function of the form $f(x;\theta)$ or $p(x;\theta)$ , where $\theta \in \Omega$ for a specified set $\Omega$ . For example, in (a) above, $\Omega = \{\theta \mid \theta > 0\}$ . We call $\theta$ a parameter of the distribution. Because $\theta$ is unknown, we want to estimate it.

Suppose, in particular, we have a random variable $X$ that has a distribution dependent on a parameter $\theta$ . Let $X_{1},X_{2},\ldots,X_{n}$ be a random sample of size $n$ from this distribution; that is, the $X_{i}$ are iid (independent and identically distributed). A statistic is a function $T = g(X_{1},X_{2},\ldots,X_{n})$ of the sample that does not depend on any unknown parameters. In this chapter we discuss several statistics that are used to estimate the unknown parameter $\theta$ or to test hypotheses about $\theta$ .

Let $X_{1},X_{2},\ldots,X_{n}$ denote a random sample of size $n$ from a distribution that has mean $\mu = E(X_{i})$ and variance $\sigma^{2} = \operatorname{Var}(X_{i})$ . The sample mean is defined by

\begin{equation}\label{eq:sample-mean}
\bar{X} = \frac{1}{n}\sum_{i=1}^{n} X_{i}.
\end{equation}

The sample mean has the following properties:

\[
E(\bar{X}) = \mu, \quad \operatorname{Var}(\bar{X}) = \frac{\sigma^{2}}{n}.
\]

By the Weak Law of Large Numbers, $\bar{X}$ converges in probability to $\mu$ as $n \to \infty$ . By the Central Limit Theorem, for large $n$ ,

\[
\frac{\bar{X}-\mu}{\sigma/\sqrt{n}} \xrightarrow{d} N(0,1).
\]

The sample variance is defined by

\begin{equation}\label{eq:sample-variance}
S^{2} = \frac{1}{n-1}\sum_{i=1}^{n} (X_{i}-\bar{X})^{2}.
\end{equation}

An unbiased estimator of $\sigma^{2}$ is $S^{2}$ because $E(S^{2}) = \sigma^{2}$ .

Other important statistics include the sample standard deviation $S = \sqrt{S^{2}}$ , the sample median, and the sample standard deviation.

\subsection{Point Estimators}\label{sub:point-estimators}

A point estimator $\hat{\theta}$ of a parameter $\theta$ is a statistic that provides a single value as an estimate of $\theta$ .

Several criteria are used to evaluate estimators:

\begin{definition}[Bias]\label{def:bias}
The bias of an estimator $\hat{\theta}$ is defined as $E(\hat{\theta}) - \theta$ . If $E(\hat{\theta}) = \theta$ , then $\hat{\theta}$ is unbiased.
\end{definition}

\begin{definition}[Mean Squared Error]\label{def:mse}
The mean squared error of an estimator $\hat{\theta}$ is $\operatorname{MSE}(\hat{\theta}) = E[(\hat{\theta}-\theta)^{2}]$ . For unbiased estimators, $\operatorname{MSE} = \operatorname{Var}(\hat{\theta})$ .
\end{definition}

\begin{example}\label{ex:sample-mean-properties}
Let $X_{1},X_{2},\ldots,X_{n}$ be a random sample from a distribution with mean $\mu$ and variance $\sigma^{2}$ . Then $\bar{X}$ is an unbiased estimator of $\mu$ because $E(\bar{X}) = \mu$ . The variance of $\bar{X}$ is $\sigma^{2}/n$ . The mean squared error of $\bar{X}$ is $\sigma^{2}/n$ .
\end{example}

\begin{theorem}[Method of Moment Matching]\label{ex:method-of-moments}
The $k$th sample moment is $M_{k} = \frac{1}{n}\sum_{i=1}^{n} X_{i}^{k}$ . The method of moments equates sample moments to population moments to obtain estimators.
\end{theorem}

\begin{example}\label{ex:normal-mle}
Let $X_{1},X_{2},\ldots,X_{n}$ be a random sample from $N(\mu,\sigma^{2})$ . The maximum likelihood estimators (MLEs) are
\[
\hat{\mu} = \bar{X}, \quad \hat{\sigma}^{2} = \frac{1}{n}\sum_{i=1}^{n} (X_{i}-\bar{X})^{2}.
\]
Note that $\hat{\mu}$ is unbiased, but $\hat{\sigma}^{2}$ is biased (though asymptotically unbiased).
\end{example}

\section{Confidence Intervals}\label{sec:confidence-intervals}

A confidence interval provides a range of plausible values for an unknown parameter.

\begin{definition}[Confidence Interval]\label{def:confidence-interval}
Let $\theta$ be an unknown parameter. Let $L = g_{1}(X_{1},\ldots,X_{n})$ and $U = g_{2}(X_{1},\ldots,X_{n})$ be two statistics such that $L \leq U$ . If for all $\theta \in \Omega$ ,
\[
P_{\theta}(L \leq \theta \leq U) = 1-\alpha,
\]
then the random interval $[L,U]$ is called a $100(1-\alpha)\%$ confidence interval for $\theta$ .
\end{definition}

The quantity $1-\alpha$ is called the confidence coefficient, and $\alpha$ is the significance level.

\begin{example}[Confidence interval for normal mean, known variance]\label{ex:ci-normal-known-variance}
Let $X_{1},X_{2},\ldots,X_{n}$ be a random sample from $N(\mu,\sigma^{2})$ where $\sigma^{2}$ is known. Then a $100(1-\alpha)\%$ confidence interval for $\mu$ is
\[
\left[ \bar{X} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}}, \quad \bar{X} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}} \right],
\]
where $z_{\alpha/2}$ satisfies $\Phi(z_{\alpha/2}) = 1-\alpha/2$ .
\end{example}

\begin{example}[Confidence interval for normal mean, unknown variance]\label{ex:ci-normal-unknown-variance}
Let $X_{1},X_{2},\ldots,X_{n}$ be a random sample from $N(\mu,\sigma^{2})$ where both $\mu$ and $\sigma^{2}$ are unknown. Then a $100(1-\alpha)\%$ confidence interval for $\mu$ is
\[
\left[ \bar{X} - t_{\alpha/2,n-1}\frac{S}{\sqrt{n}}, \quad \bar{X} + t_{\alpha/2,n-1}\frac{S}{\sqrt{n}} \right],
\]
where $t_{\alpha/2,n-1}$ is the upper $\alpha/2$ critical value of the $t(n-1)$ distribution.
\end{example}

\begin{theorem}[Confidence interval for variance]\label{ex:ci-variance}
Let $X_{1},X_{2},\ldots,X_{n}$ be a random sample from $N(\mu,\sigma^{2})$ . A $100(1-\alpha)\%$ confidence interval for $\sigma^{2}$ is
\[
\left[ \frac{(n-1)S^{2}}{\chi^{2}_{\alpha/2,n-1}}, \quad \frac{(n-1)S^{2}}{\chi^{2}_{1-\alpha/2,n-1}} \right],
\]
where $\chi^{2}_{\alpha/2,n-1}$ and $\chi^{2}_{1-\alpha/2,n-1}$ are critical values of the $\chi^{2}(n-1)$ distribution.
\end{theorem}

\subsection{Confidence Intervals for Difference in Means}\label{sub:ci-difference-means}

Let $X_{1},\ldots,X_{n_{1}}$ be a random sample from $N(\mu_{1},\sigma_{1}^{2})$ and $Y_{1},\ldots,Y_{n_{2}}$ be an independent random sample from $N(\mu_{2},\sigma_{2}^{2})$ .

If $\sigma_{1}^{2}$ and $\sigma_{2}^{2}$ are known, a $100(1-\alpha)\%$ confidence interval for $\mu_{1}-\mu_{2}$ is
\[
\left[ \bar{X}-\bar{Y} \pm z_{\alpha/2}\sqrt{\frac{\sigma_{1}^{2}}{n_{1}}+\frac{\sigma_{2}^{2}}{n_{2}}} \right].
\]

If $\sigma_{1}^{2}=\sigma_{2}^{2}=\sigma^{2}$ but unknown, the pooled estimator is
\[
S_{p}^{2} = \frac{(n_{1}-1)S_{1}^{2}+(n_{2}-1)S_{2}^{2}}{n_{1}+n_{2}-2}.
\]
A $100(1-\alpha)\%$ confidence interval is
\[
\left[ \bar{X}-\bar{Y} \pm t_{\alpha/2,n_{1}+n_{2}-2}S_{p}\sqrt{\frac{1}{n_{1}}+\frac{1}{n_{2}}} \right].
\]

\subsection{Confidence Interval for Difference in Proportions}\label{sub:ci-difference-proportions}

Let $X \sim \operatorname{Bin}(n_{1},p_{1})$ and $Y \sim \operatorname{Bin}(n_{2},p_{2})$ be independent. Large-sample approximations yield the following $100(1-\alpha)\%$ confidence interval for $p_{1}-p_{2}$ :
\[
\left[ \hat{p}_{1}-\hat{p}_{2} \pm z_{\alpha/2}\sqrt{\frac{\hat{p}_{1}(1-\hat{p}_{1})}{n_{1}}+\frac{\hat{p}_{2}(1-\hat{p}_{2})}{n_{2}}} \right],
\]
where $\hat{p}_{1}=X/n_{1}$ and $\hat{p}_{2}=Y/n_{2}$ .

\section{Order Statistics}\label{sec:order-statistics}

Let $X_{1},X_{2},\ldots,X_{n}$ be a random sample from a distribution with cdf $F(x)$ . Define the order statistics by arranging the sample in ascending order:

\[
X_{(1)} \leq X_{(2)} \leq \cdots \leq X_{(n)}.
\]

The $k$th order statistic is $X_{(k)}$ . The sample median is $X_{((n+1)/2)}$ if $n$ is odd, or the average of the two middle values if $n$ is even.

The sample range is $R = X_{(n)} - X_{(1)}$ and the sample interquartile range is $IQR = X_{(3n/4)} - X_{(n/4)}$ (approximately).

The cdf and pdf of the $k$th order statistic $X_{(k)}$ are:

\[
F_{X_{(k)}}(x) = \sum_{j=k}^{n} \binom{n}{j} [F(x)]^{j} [1-F(x)]^{n-j},
\]

\[
f_{X_{(k)}}(x) = \frac{n!}{(k-1)!(n-k)!} [F(x)]^{k-1} [1-F(x)]^{n-k} f(x).
\]

\subsection{Quantiles}\label{sub:quantiles}

The population quantile $x_{p}$ is defined by $F(x_{p}) = p$ .

\begin{definition}[Sample Quantile]\label{def:sample-quantile}
The sample quantile $x_{p}$ can be estimated by the order statistic $X_{(\lceil np \rceil)}$ or by linear interpolation.
\end{definition}

Common quantiles include:
\begin{itemize}
\item Median: $p = 0.5$
\item Quartiles: $p = 0.25, 0.75$
\item Deciles: $p = 0.1, 0.2, \ldots, 0.9$
\item Percentiles: $p = 0.01, 0.02, \ldots, 0.99$
\end{itemize}

\section{Introduction to Hypothesis Testing}\label{sec:hypothesis-testing}

A statistical hypothesis is a statement about the distribution of a random variable. We test a null hypothesis $H_{0}$ against an alternative hypothesis $H_{1}$ .

\begin{definition}[Null and Alternative Hypotheses]\label{def:hypotheses}
The null hypothesis $H_{0}$ is the hypothesis being tested. The alternative hypothesis $H_{1}$ is the hypothesis we accept if $H_{0}$ is rejected.
\end{definition}

The four possible outcomes of a hypothesis test are:

\begin{center}
\begin{tabular}{c|c|c}
 & Fail to Reject $H_{0}$ & Reject $H_{0}$ \\ \hline
$H_{0}$ True & Correct & Type I Error ($\alpha$) \\
$H_{0}$ False & Type II Error ($\beta$) & Correct Power ($1-\beta$)
\end{tabular}
\end{center}

The significance level $\alpha = P(\text{Type I Error}) = P(\text{Reject } H_{0} \mid H_{0} \text{ true})$ .

\begin{example}[Test for normal mean, known variance]\label{ex:test-normal-known}
Let $X_{1},\ldots,X_{n} \sim N(\mu,\sigma^{2})$ with known $\sigma^{2}$ . To test $H_{0}: \mu = \mu_{0}$ against $H_{1}: \mu > \mu_{0}$ , we use the test statistic
\[
Z = \frac{\bar{X}-\mu_{0}}{\sigma/\sqrt{n}}.
\]
We reject $H_{0}$ at level $\alpha$ if $Z > z_{\alpha}$ .
\end{example}

\begin{example}[Test for normal mean, unknown variance]\label{ex:test-normal-unknown}
Let $X_{1},\ldots,X_{n} \sim N(\mu,\sigma^{2})$ with unknown $\sigma^{2}$ . To test $H_{0}: \mu = \mu_{0}$ against $H_{1}: \mu \neq \mu_{0}$ , we use the test statistic
\[
T = \frac{\bar{X}-\mu_{0}}{S/\sqrt{n}}.
\]
We reject $H_{0}$ at level $\alpha$ if $|T| > t_{\alpha/2,n-1}$ .
\end{example}

\section{Chi-Square Tests}\label{sec:chi-square-tests}

The chi-square test is used for categorical data.

\subsection{Goodness-of-Fit Test}\label{sub:goodness-of-fit}

Let $X_{1},\ldots,X_{n}$ be iid observations from a categorical distribution with $k$ categories. Under $H_{0}$, the probabilities are $p_{1,0},p_{2,0},\ldots,p_{k,0}$ . The observed counts are $O_{1},\ldots,O_{k}$ and the expected counts are $E_{i} = np_{i,0}$ .

The chi-square goodness-of-fit statistic is

\[
\chi^{2} = \sum_{i=1}^{k} \frac{(O_{i}-E_{i})^{2}}{E_{i}}.
\]

Under $H_{0}$ , this statistic has approximately a $\chi^{2}(k-1)$ distribution for large $n$ . We reject $H_{0}$ if $\chi^{2} > \chi^{2}_{\alpha,k-1}$ .

\subsection{Test of Independence}\label{sub:test-independence}

For a contingency table with $r$ rows and $c$ columns, let $O_{ij}$ be the observed count in cell $(i,j)$ . The expected count under independence is $E_{ij} = (O_{i\cdot}O_{\cdot j})/n$ .

The chi-square test statistic for independence is

\[
\chi^{2} = \sum_{i=1}^{r}\sum_{j=1}^{c} \frac{(O_{ij}-E_{ij})^{2}}{E_{ij}}.
\]

Under $H_{0}$ of independence, $\chi^{2} \approx \chi^{2}((r-1)(c-1))$ .

\section{The Method of Monte Carlo}\label{sec:monte-carlo}

Monte Carlo methods use random sampling to approximate statistical quantities.

The law of large numbers states that if $E[g(X)] = \mu$ exists, then

\[
\frac{1}{n}\sum_{i=1}^{n} g(X_{i}) \xrightarrow{P} \mu \quad \text{as } n \to \infty .
\]

Thus, we can approximate $\mu$ by simulating $n$ values from the distribution of $X$ and averaging $g(X_{i})$ .

\begin{example}[Approximating $\pi$]\label{ex:monte-carlo-pi}
We can approximate $\pi$ by simulating points in the unit square $[0,1] \times [0,1]$ and computing the proportion that fall inside the unit quarter circle. The area of the quarter circle is $\pi/4$ , so $\hat{\pi} = 4 \times \text{(proportion inside)}$ .
\end{example}

\subsection{Accept-Reject Generation Algorithm}\label{sub:accept-reject}

The accept-reject algorithm generates random samples from a target distribution $f(x)$ using a proposal distribution $g(x)$ with known density.

\begin{example}[Accept-Reject Algorithm]\label{ex:accept-reject}
\begin{enumerate}
\item Choose a proposal density $g(x)$ such that $f(x) \leq Mg(x)$ for all $x$ and some constant $M > 1$ .
\item Generate $Y \sim g$ and $U \sim \operatorname{Uniform}(0,1)$ independently.
\item If $U \leq f(Y)/(Mg(Y))$ , accept $X = Y$ ; otherwise, reject and return to step 2.
\end{enumerate}
\end{example}

The acceptance probability is $1/M$ . The algorithm generates samples from $f$ exactly.

\section{Bootstrap Procedures}\label{sec:bootstrap}

Bootstrap methods use resampling to estimate the sampling distribution of a statistic.

Let $X_{1},\ldots,X_{n}$ be a random sample and $T = g(X_{1},\ldots,X_{n})$ be a statistic. The bootstrap proceeds as follows:

\begin{enumerate}
\item Resample $n$ observations with replacement from the original sample to form a bootstrap sample $X_{1}^{*},\ldots,X_{n}^{*}$ .
\item Compute the bootstrap replicate $T^{*} = g(X_{1}^{*},\ldots,X_{n}^{*})$ .
\item Repeat steps 1-2 many times (say $B$ times) to obtain $T_{1}^{*},\ldots,T_{B}^{*}$ .
\item Approximate the distribution of $T$ by the empirical distribution of $T_{1}^{*},\ldots,T_{B}^{*}$ .
\end{enumerate}

\subsection{Percentile Bootstrap Confidence Intervals}\label{sub:percentile-bootstrap}

A $100(1-\alpha)\%$ percentile bootstrap confidence interval for $\theta$ is

\[
\left[ T_{(\alpha/2)}^{*}, T_{(1-\alpha/2)}^{*} \right],
\]

where $T_{(k)}^{*}$ is the $k$th order statistic of the bootstrap replicates.

\section*{Exercises}\label{sec:chapter4-exercises}

\begin{exercise}{4.1.1}
Let $X_{1},X_{2},X_{3},X_{4},X_{5}$ be a random sample of size 5 from a distribution with mean $\mu = 10$ and variance $\sigma^{2} = 4$ . Find the mean and variance of $\bar{X}$ .
\end{exercise}

\begin{exercise}{4.1.3}
Let $X_{1},\ldots,X_{n}$ be iid with mean $\mu$ and variance $\sigma^{2}$ . Show that $S^{2}$ is an unbiased estimator of $\sigma^{2}$ .
\end{exercise}

\begin{exercise}{4.2.1}
Let $X_{1},\ldots,X_{16}$ be a random sample from $N(50,25)$ . Find $P(48 < \bar{X} < 52)$ .
\end{exercise}

\begin{exercise}{4.2.3}
Construct a $95\%$ confidence interval for the mean of a normal distribution based on the following sample: $n=25$ , $\bar{x}=100$ , $s=15$ .
\end{exercise}

\begin{exercise}{4.3.1}
Two independent samples are given: Sample 1: $n_{1}=30$ , $\bar{x}_{1}=75$ , $s_{1}=10$ ; Sample 2: $n_{2}=40$ , $\bar{x}_{2}=80$ , $s_{2}=12$ . Assuming equal variances, find a $90\%$ confidence interval for $\mu_{1}-\mu_{2}$ .
\end{exercise}

\begin{exercise}{4.4.1}
Let $X_{1},\ldots,X_{10}$ be a random sample from $N(0,1)$ . Find the probability that the sample median exceeds 0.5.
\end{exercise}

\begin{exercise}{4.5.1}
A manufacturer claims that the mean lifetime of a light bulb is at least 1000 hours. A sample of 25 bulbs has a mean lifetime of 950 hours with a standard deviation of 80 hours. Test the claim at the $\alpha=0.05$ level.
\end{exercise}

\begin{exercise}{4.7.1}
A chi-square goodness-of-fit test is performed on a die to test whether it is fair. The die is rolled 120 times with the following frequencies: Face 1: 18, Face 2: 22, Face 3: 16, Face 4: 24, Face 5: 19, Face 6: 21. Compute the test statistic and $p$-value.
\end{exercise}

\begin{exercise}{4.8.1}
Describe the accept-reject algorithm for generating random samples from a gamma distribution using an exponential proposal distribution.
\end{exercise}

\begin{exercise}{4.9.1}
Explain the bootstrap procedure for estimating the standard error of the sample median.
\end{exercise}

\userannotation{Chapter 4 covers statistical inference: sampling, point estimation, confidence intervals, hypothesis testing, chi-square tests, Monte Carlo methods, and bootstrap procedures. Key concepts: sample mean and variance, unbiased estimation, CI formulas for normal means and variances, hypothesis testing framework, Type I/II errors, bootstrap resampling.}
