\chapter{Some Special Distributions}\label{chap:special-distributions}

\section{The Binomial and Related Distributions}\label{sec:binomial-distributions}

In Chapter 1 we introduced the uniform distribution and the hypergeometric distribution. In this chapter we discuss some other important distributions of random variables frequently used in statistics. We begin with the binomial and related distributions.

A Bernoulli experiment is a random experiment, the outcome of which can be classified in but one of two mutually exclusive and exhaustive ways, for instance, success or failure (e.g., female or male, life or death, nondefective or defective). A sequence of Bernoulli trials occurs when a Bernoulli experiment is performed several independent times so that the probability of success, say $p$ , remains the same from trial to trial.

Let $X$ be a random variable associated with a Bernoulli trial by defining it as follows:

\[
X(\text{success}) = 1 \quad \text{and} \quad X(\text{failure}) = 0.
\]

That is, the two outcomes, success and failure, are denoted by one and zero, respectively. The pmf of $X$ can be written as

\begin{equation}\label{eq:bernoulli-pmf}
p(x) = p^{x}(1-p)^{1-x}, \quad x=0,1,
\end{equation}

and we say that $X$ has a Bernoulli distribution. The expected value of $X$ is

\[
\mu = E(X) = (0)(1-p) + (1)(p) = p,
\]

and the variance of $X$ is

\[
\sigma^{2} = \operatorname{Var}(X) = p^{2}(1-p) + (1-p)^{2}p = p(1-p).
\]

It follows that the standard deviation of $X$ is $\sigma = \sqrt{p(1-p)}$ .

In a sequence of $n$ independent Bernoulli trials, where the probability of success remains constant, let $X_{i}$ denote the Bernoulli random variable associated with the $i$th trial. In such a sequence of Bernoulli trials, we are often interested in the total number of successes and not in the order of their occurrence. If we let the random variable $X$ equal the number of observed successes in $n$ Bernoulli trials, the possible values of $X$ are $0,1,2,\ldots,n$ . If $x$ successes occur, where $x=0,1,2,\ldots,n$ , then $n-x$ failures occur. The number of ways of selecting the $x$ positions for the $x$ successes in the $n$ trials is

\[
\binom{n}{x} = \frac{n!}{x!(n-x)!}.
\]

Since the trials are independent and the probabilities of success and failure on each trial are, respectively, $p$ and $1-p$ , the probability of each of these ways is $p^{x}(1-p)^{n-x}$ . Thus the pmf of $X$ is

\begin{equation}\label{eq:binomial-pmf}
p(x) = \begin{cases}
\binom{n}{x} p^{x}(1-p)^{n-x} & x=0,1,2,\ldots,n \\
0 & \text{elsewhere}.
\end{cases}
\end{equation}

A random variable $X$ that has a pmf of the form of $p(x)$ is said to have a binomial distribution, and any such $p(x)$ is called a binomial pmf. A binomial distribution is denoted by the symbol $b(n,p)$ . The constants $n$ and $p$ are called the parameters of the binomial distribution.

\begin{example}\label{ex:binomial-dice}
Suppose we roll a fair six-sided die 3 times. What is the probability of getting exactly 2 sixes? For our notation, let $X$ be the number of sixes obtained in the 3 rolls. Then $X$ has a binomial distribution with $n=3$ and $p=1/6$ . Hence,

\[
P(X=2) = p(2) = \binom{3}{2} \left(\frac{1}{6}\right)^{2} \left(\frac{5}{6}\right)^{1} = 0.06944.
\]
\end{example}

The mgf of a binomial distribution is easily obtained as follows:

\[
M(t) = \sum_{x} e^{tx} p(x) = \sum_{x=0}^{n} \binom{n}{x} (pe^{t})^{x} (1-p)^{n-x} = [(1-p) + pe^{t}]^{n}.
\]

The mean $\mu$ and the variance $\sigma^{2}$ of $X$ are

\[
\mu = np, \quad \sigma^{2} = np(1-p).
\]

\begin{theorem}\label{ex:binomial-sum}
Suppose $X_{1},X_{2},\ldots,X_{m}$ are independent random variables such that $X_{i}$ has binomial $b(n_{i},p)$ distribution, for $i=1,2,\ldots,m$ . Let $Y = \sum_{i=1}^{m} X_{i}$ . Then $Y$ has a binomial $b(\sum_{i=1}^{m} n_{i},p)$ distribution.
\end{theorem}

This result follows from the mgf or by convolution of pmfs.

\subsection{Negative Binomial and Geometric Distributions}\label{sub:negative-binomial}

Consider a sequence of independent Bernoulli trials with constant probability $p$ of success. Let the random variable $Y$ denote the total number of failures in this sequence before the $r$th success, that is, $Y+r$ is equal to the number of trials necessary to produce exactly $r$ successes with the last trial as a success. Here $r$ is a fixed positive integer. The pmf of $Y$ is

\begin{equation}\label{eq:negative-binomial-pmf}
p_{Y}(y) = \begin{cases}
\binom{y+r-1}{r-1} p^{r}(1-p)^{y} & y=0,1,2,\ldots \\
0 & \text{elsewhere}.
\end{cases}
\end{equation}

A distribution with a pmf of the form $p_{Y}(y)$ is called a negative binomial distribution.

If $r=1$ , then $Y$ has the pmf

\begin{equation}\label{eq:geometric-pmf}
p_{Y}(y) = p(1-p)^{y}, \quad y=0,1,2,\ldots,
\end{equation}

zero elsewhere. In this special case, $r=1$ , we say that $Y$ has a geometric distribution. In terms of Bernoulli trials, $Y$ is the number of failures until the first success.

The geometric distribution has the memoryless property:

\[
P(Y \geq k+j \mid Y \geq k) = P(Y \geq j).
\]

\subsection{Multinomial Distribution}\label{sub:multinomial-distribution}

The binomial distribution is generalized to the multinomial distribution as follows. Let a random experiment be repeated $n$ independent times. On each repetition, there is one and only one outcome from one of $k$ categories. Call the categories $C_{1},C_{2},\ldots,C_{k}$ . For each $i=1,\ldots,k$ , let $p_{i}$ be the probability that the outcome is an element of $C_{i}$ and assume that $p_{i}$ remains constant throughout the $n$ independent repetitions. Define the random variable $X_{i}$ to be equal to the number of outcomes that are elements of $C_{i}$ , $i=1,2,\ldots,k-1$ . Because $X_{k} = n-X_{1}-\cdots-X_{k-1}$ , $X_{k}$ is determined by the other $X_{i}$'s.

The joint pmf of $(X_{1},X_{2},\ldots,X_{k-1})$ is

\begin{equation}\label{eq:multinomial-pmf}
P(X_{1}=x_{1},X_{2}=x_{2},\ldots,X_{k-1}=x_{k-1}) = \frac{n!}{x_{1}!\cdots x_{k-1}!x_{k}!} p_{1}^{x_{1}}\cdots p_{k-1}^{x_{k-1}} p_{k}^{x_{k}},
\end{equation}

for all $x_{1},x_{2},\ldots,x_{k-1}$ that are nonnegative integers and such that $x_{1}+x_{2}+\cdots+x_{k-1} \leq n$ , where $x_{k} = n-x_{1}-\cdots-x_{k-1}$ and $p_{k} = 1-\sum_{j=1}^{k-1} p_{j}$ .

We say that $(X_{1},X_{2},\ldots,X_{k-1})$ has a multinomial distribution with parameters $n$ and $p_{1},\ldots,p_{k-1}$ .

\subsection{Hypergeometric Distribution}\label{sub:hypergeometric}

Suppose we have a lot of $N$ items of which $D$ are defective. Let $X$ denote the number of defective items in a sample of size $n$ . If the sampling is done without replacement, the pmf of $X$ is

\begin{equation}\label{eq:hypergeometric-pmf}
p(x) = \frac{\binom{N-D}{n-x}\binom{D}{x}}{\binom{N}{n}}, \quad x=0,1,\ldots,n,
\end{equation}

where, as usual, a binomial coefficient is taken to be 0 when the top value is less than the bottom value. We say that $X$ has a hypergeometric distribution with parameters $(N,D,n)$ .

The mean of $X$ is

\[
E(X) = n\frac{D}{N},
\]

and the variance is

\begin{equation}\label{eq:hypergeometric-variance}
\operatorname{Var}(X) = n\frac{D}{N}\frac{N-D}{N}\frac{N-n}{N-1}.
\end{equation}

The last term is often thought of as the correction term when sampling without replacement. Note that it is close to 1 if $N$ is much larger than $n$ .

\section{The Poisson Distribution}\label{sec:poisson-distribution}

Recall that the following series expansion holds for all real numbers $z$ :

\[
e^{z} = \sum_{x=0}^{\infty} \frac{z^{x}}{x!}.
\]

Consider the function $p(x)$ defined by

\begin{equation}\label{eq:poisson-pmf}
p(x) = \begin{cases}
\frac{\lambda^{x}e^{-\lambda}}{x!} & x=0,1,2,\ldots \\
0 & \text{elsewhere},
\end{cases}
\end{equation}

where $\lambda > 0$ . Since $\lambda > 0$ , then $p(x) \geq 0$ and

\[
\sum_{x=0}^{\infty} p(x) = \sum_{x=0}^{\infty} \frac{\lambda^{x}e^{-\lambda}}{x!} = e^{-\lambda} \sum_{x=0}^{\infty} \frac{\lambda^{x}}{x!} = e^{-\lambda}e^{\lambda} = 1;
\]

that is, $p(x)$ satisfies the conditions of being a pmf of a discrete type of random variable. A random variable that has a pmf of the form $p(x)$ is said to have a Poisson distribution with parameter $\lambda$ .

The Poisson distribution occurs in many areas of applications. Consider a process that counts the number of certain events occurring over an interval of time; for example, the number of tornados that touch down in Michigan per year, the number of cars entering a parking lot between 8:00 and 12:00 on a weekday, the number of car accidents at a busy intersection per week, the number of typographical errors per page of a manuscript, and the number of blemishes on a manufactured car door.

The mean and variance of a Poisson distribution are both $\lambda$ :

\[
E(X) = \lambda, \quad \operatorname{Var}(X) = \lambda.
\]

The mgf of a Poisson distribution is

\[
M(t) = \exp\{\lambda(e^{t}-1)\}.
\]

\begin{theorem}\label{ex:poisson-approximation}
If $X_{n} \sim \operatorname{Bin}(n,p_{n})$ and $np_{n} \to \lambda$ as $n \to \infty$ , then $X_{n} \xrightarrow{d} \operatorname{Poisson}(\lambda)$ .
\end{theorem}

This theorem shows that the Poisson distribution can be used as an approximation to the binomial distribution when $n$ is large and $p$ is small.

\section{The $\Gamma$, $\chi^{2}$, and $\beta$ Distributions}\label{sec:gamma-chi2-beta}

In this section we discuss several continuous distributions that are related to the gamma function.

\subsection{The Gamma Distribution}\label{sub:gamma-distribution}

The gamma function is defined by

\[
\Gamma(\alpha) = \int_{0}^{\infty} x^{\alpha-1}e^{-x} \, dx, \quad \alpha > 0.
\]

The gamma function satisfies the recursion $\Gamma(\alpha+1) = \alpha\Gamma(\alpha)$ , and for positive integers, $\Gamma(n) = (n-1)!$ .

\begin{definition}[Gamma Distribution]\label{def:gamma-distribution}
A random variable $X$ is said to have a gamma distribution with parameters $\alpha > 0$ and $\beta > 0$ if its pdf is

\[
f(x) = \begin{cases}
\frac{1}{\Gamma(\alpha)\beta^{\alpha}} x^{\alpha-1} e^{-x/\beta} & x > 0 \\
0 & \text{elsewhere}.
\end{cases}
\]

We write $X \sim \Gamma(\alpha,\beta)$ .
\end{definition}

The parameters $\alpha$ and $\beta$ are called the shape and scale parameters, respectively. The mean and variance of $X$ are

\[
E(X) = \alpha\beta, \quad \operatorname{Var}(X) = \alpha\beta^{2}.
\]

The mgf of a gamma distribution is

\[
M(t) = (1-\beta t)^{-\alpha}, \quad t < \frac{1}{\beta}.
\]

Two important special cases of the gamma distribution are:
\begin{itemize}
\item $\Gamma(1,\beta)$ is the exponential distribution $\operatorname{Exp}(1/\beta)$ .
\item $\Gamma(\alpha,2)$ with $\alpha = n/2$ is related to the chi-square distribution.
\end{itemize}

If $X_{1} \sim \Gamma(\alpha_{1},\beta)$ and $X_{2} \sim \Gamma(\alpha_{2},\beta)$ are independent, then $X_{1}+X_{2} \sim \Gamma(\alpha_{1}+\alpha_{2},\beta)$ .

\subsection{The $\chi^{2}$-Distribution}\label{sub:chi-square-distribution}

\begin{definition}[Chi-Square Distribution]\label{def:chi-square}
Let $Z_{1},Z_{2},\ldots,Z_{n}$ be iid $N(0,1)$ random variables. Then the random variable

\[
X = \sum_{i=1}^{n} Z_{i}^{2}
\]

is said to have a chi-square distribution with $n$ degrees of freedom. We write $X \sim \chi^{2}(n)$ .
\end{definition}

The chi-square distribution is a special case of the gamma distribution with $\alpha = n/2$ and $\beta = 2$ . That is, $\chi^{2}(n) \equiv \Gamma(n/2,2)$ .

The pdf of $X \sim \chi^{2}(n)$ is

\[
f(x) = \begin{cases}
\frac{1}{\Gamma(n/2)2^{n/2}} x^{n/2-1} e^{-x/2} & x > 0 \\
0 & \text{elsewhere}.
\end{cases}
\]

The mean and variance of a chi-square distribution are

\[
E(X) = n, \quad \operatorname{Var}(X) = 2n.
\]

The chi-square distribution has the additive property: if $X_{1} \sim \chi^{2}(m)$ and $X_{2} \sim \chi^{2}(n)$ are independent, then $X_{1}+X_{2} \sim \chi^{2}(m+n)$ .

\subsection{The $\beta$-Distribution}\label{sub:beta-distribution}

\begin{definition}[Beta Distribution]\label{def:beta-distribution}
A random variable $X$ is said to have a beta distribution with parameters $\alpha > 0$ and $\beta > 0$ if its pdf is

\[
f(x) = \begin{cases}
\frac{1}{B(\alpha,\beta)} x^{\alpha-1}(1-x)^{\beta-1} & 0 < x < 1 \\
0 & \text{elsewhere},
\end{cases}
\]

where $B(\alpha,\beta) = \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha+\beta)}$ is the beta function.
\end{definition}

The mean and variance of $X \sim \operatorname{Beta}(\alpha,\beta)$ are

\[
E(X) = \frac{\alpha}{\alpha+\beta}, \quad \operatorname{Var}(X) = \frac{\alpha\beta}{(\alpha+\beta)^{2}(\alpha+\beta+1)}.
\]

Note that $\operatorname{Beta}(1,1)$ is the uniform distribution on $(0,1)$ .

\section{The Normal Distribution}\label{sec:normal-distribution}

The normal distribution is one of the most important distributions in statistics.

\begin{definition}[Normal Distribution]\label{def:normal-distribution}
A random variable $X$ is said to have a normal distribution with parameters $\mu \in \mathbb{R}$ and $\sigma^{2} > 0$ if its pdf is

\[
f(x) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left\{ -\frac{(x-\mu)^{2}}{2\sigma^{2}} \right\}, \quad -\infty < x < \infty.
\]

We write $X \sim N(\mu,\sigma^{2})$ .
\end{definition}

When $\mu = 0$ and $\sigma^{2} = 1$ , we say that $X$ has a standard normal distribution, denoted by $Z \sim N(0,1)$ . The pdf of $Z$ is

\[
\phi(z) = \frac{1}{\sqrt{2\pi}} e^{-z^{2}/2}, \quad -\infty < z < \infty,
\]

and the cdf is $\Phi(z) = P(Z \leq z)$ .

If $X \sim N(\mu,\sigma^{2})$ , then the standardized variable $Z = (X-\mu)/\sigma \sim N(0,1)$ . This property is extremely useful for computing probabilities:

\[
P(X \leq x) = P\!\left(Z \leq \frac{x-\mu}{\sigma}\right) = \Phi\!\left(\frac{x-\mu}{\sigma}\right).
\]

The mean and variance of $X \sim N(\mu,\sigma^{2})$ are $E(X) = \mu$ and $\operatorname{Var}(X) = \sigma^{2}$ .

The mgf of $X$ is

\[
M(t) = \exp\left\{\mu t + \frac{\sigma^{2}t^{2}}{2}\right\}.
\]

\begin{theorem}[Linear Combinations of Normals]\label{ex:normal-linear-combination}
If $X_{1} \sim N(\mu_{1},\sigma_{1}^{2})$ and $X_{2} \sim N(\mu_{2},\sigma_{2}^{2})$ are independent, then for any constants $a$ and $b$ ,

\[
aX_{1} + bX_{2} \sim N(a\mu_{1}+b\mu_{2}, a^{2}\sigma_{1}^{2}+b^{2}\sigma_{2}^{2}).
\]

More generally, any linear combination of independent normal random variables is normally distributed.
\end{theorem}

\section{The Multivariate Normal Distribution}\label{sec:multivariate-normal}

\subsection{Bivariate Normal Distribution}\label{sub:bivariate-normal-ch3}

We say that $(X_{1},X_{2})$ has a bivariate normal distribution if the joint pdf is given by

\[
f(x_{1},x_{2}) = \frac{1}{2\pi\sigma_{1}\sigma_{2}\sqrt{1-\rho^{2}}} \exp\left\{ -\frac{1}{2(1-\rho^{2})} \left[ \frac{(x_{1}-\mu_{1})^{2}}{\sigma_{1}^{2}} - 2\rho\frac{(x_{1}-\mu_{1})(x_{2}-\mu_{2})}{\sigma_{1}\sigma_{2}} + \frac{(x_{2}-\mu_{2})^{2}}{\sigma_{2}^{2}} \right] \right\},
\]

for $-\infty < x_{1} < \infty$ , $-\infty < x_{2} < \infty$ , where $\mu_{1} \in \mathbb{R}$ , $\mu_{2} \in \mathbb{R}$ , $\sigma_{1} > 0$ , $\sigma_{2} > 0$ , and $-1 < \rho < 1$ . We write $(X_{1},X_{2}) \sim N(\mu_{1},\mu_{2},\sigma_{1}^{2},\sigma_{2}^{2},\rho)$ .

The parameters have the following interpretations:
\begin{itemize}
\item $E(X_{1}) = \mu_{1}$ and $E(X_{2}) = \mu_{2}$ .
\item $\operatorname{Var}(X_{1}) = \sigma_{1}^{2}$ and $\operatorname{Var}(X_{2}) = \sigma_{2}^{2}$ .
\item $\rho$ is the correlation coefficient between $X_{1}$ and $X_{2}$ .
\end{itemize}

\begin{theorem}[Properties of Bivariate Normal]\label{ex:bivariate-normal-properties-ch3}
If $(X_{1},X_{2}) \sim N(\mu_{1},\mu_{2},\sigma_{1}^{2},\sigma_{2}^{2},\rho)$ , then:

\begin{enumerate}
\item The marginal distributions are $X_{1} \sim N(\mu_{1},\sigma_{1}^{2})$ and $X_{2} \sim N(\mu_{2},\sigma_{2}^{2})$ .
\item The conditional distribution of $X_{2}$ given $X_{1}=x_{1}$ is normal with mean and variance:
\[
E(X_{2}\mid x_{1}) = \mu_{2} + \rho\frac{\sigma_{2}}{\sigma_{1}}(x_{1}-\mu_{1}), \quad \operatorname{Var}(X_{2}\mid x_{1}) = \sigma_{2}^{2}(1-\rho^{2}).
\]
\item $X_{1}$ and $X_{2}$ are independent if and only if $\rho = 0$ .
\item Any linear combination $aX_{1}+bX_{2}$ has a normal distribution.
\end{enumerate}
\end{theorem}

\section{$t$- and $F$-Distributions}\label{sec:t-f-distributions}

\subsection{The $t$-Distribution}\label{sub:t-distribution-ch3}

\begin{definition}[$t$-Distribution]\label{def:t-distribution-ch3}
Let $Z \sim N(0,1)$ and $X \sim \chi^{2}(n)$ be independent. Then the random variable

\[
T = \frac{Z}{\sqrt{X/n}}
\]

is said to have a $t$-distribution with $n$ degrees of freedom. We write $T \sim t(n)$ .
\end{definition}

The pdf of $T \sim t(n)$ is

\[
f(t) = \frac{\Gamma((n+1)/2)}{\sqrt{n\pi}\,\Gamma(n/2)} \left(1+\frac{t^{2}}{n}\right)^{-(n+1)/2}, \quad -\infty < t < \infty.
\]

The mean and variance of $T$ are

\[
E(T) = 0 \quad (n > 1), \quad \operatorname{Var}(T) = \frac{n}{n-2} \quad (n > 2).
\]

As $n \to \infty$ , $t(n) \xrightarrow{d} N(0,1)$ .

When $n=1$ , the $t$-distribution is the Cauchy distribution.

\subsection{The $F$-Distribution}\label{sub:f-distribution-ch3}

\begin{definition}[$F$-Distribution]\label{def:f-distribution-ch3}
Let $X_{1} \sim \chi^{2}(m)$ and $X_{2} \sim \chi^{2}(n)$ be independent. Then the random variable

\[
F = \frac{X_{1}/m}{X_{2}/n}
\]

is said to have an $F$-distribution with degrees of freedom $m$ and $n$ . We write $F \sim F(m,n)$ .
\end{definition}

The pdf of $F \sim F(m,n)$ is

\[
f(x) = \frac{\Gamma((m+n)/2)}{\Gamma(m/2)\Gamma(n/2)} \left(\frac{m}{n}\right)^{m/2} x^{m/2-1} \left(1+\frac{mx}{n}\right)^{-(m+n)/2}, \quad x > 0.
\]

The mean and variance of $F$ are

\[
E(F) = \frac{n}{n-2} \quad (n > 2), \quad \operatorname{Var}(F) = \frac{2n^{2}(m+n-2)}{m(n-2)^{2}(n-4)} \quad (n > 4).
\]

Note that if $F \sim F(m,n)$ , then $1/F \sim F(n,m)$ .

\section*{Exercises}\label{sec:chapter3-exercises}

\begin{exercise}{3.1.1}
If the mgf of a random variable $X$ is $\left(\frac{1}{3}+\frac{2}{3}e^{t}\right)^{5}$ , find $P(X=2 \text{ or } 3)$ .
\end{exercise}

\begin{exercise}{3.1.3}
If $X$ is $b(n,p)$ , show that $E(X/n) = p$ and $E[(X/n-p)^{2}] = p(1-p)/n$ .
\end{exercise}

\begin{exercise}{3.1.11}
Toss two nickels and three dimes at random. Make appropriate assumptions and compute the probability that there are more heads showing on the nickels than on the dimes.
\end{exercise}

\begin{exercise}{3.2.1}
Let $X$ have a Poisson distribution with parameter $\lambda = 3$ . Find $P(X=2)$ and $P(X \geq 1)$ .
\end{exercise}

\begin{exercise}{3.3.1}
Show that the gamma function $\Gamma(\alpha)$ satisfies $\Gamma(\alpha+1) = \alpha\Gamma(\alpha)$ .
\end{exercise}

\begin{exercise}{3.3.3}
Let $X \sim \chi^{2}(10)$ . Find $P(X > 15.99)$ using R.
\end{exercise}

\begin{exercise}{3.4.1}
Let $X \sim N(100,225)$ (i.e., $\sigma=15$ ). Find:
\begin{enumerate}
\item $P(X < 85)$ ;
\item $P(90 < X < 115)$ ;
\item $P(X > 120)$ .
\end{enumerate}
\end{exercise}

\begin{exercise}{3.4.3}
If $X_{1} \sim N(50,36)$ and $X_{2} \sim N(30,16)$ are independent, find the distribution of $Y = 2X_{1} - 3X_{2}$ .
\end{exercise}

\begin{exercise}{3.5.1}
If $(X,Y)$ has a bivariate normal distribution with $\mu_{1}=2$ , $\mu_{2}=3$ , $\sigma_{1}^{2}=4$ , $\sigma_{2}^{2}=9$ , and $\rho=1/2$ , find $E(Y\mid X=2)$ and $\operatorname{Var}(Y\mid X=2)$ .
\end{exercise}

\begin{exercise}{3.6.1}
Let $T \sim t(10)$ . Find $P(|T| > 2.228)$ .
\end{exercise}

\begin{exercise}{3.6.5}
Let $F \sim F(5,10)$ . Find $E(F)$ (if it exists) and $P(F > 3.33)$ .
\end{exercise}

\userannotation{Chapter 3 covers special distributions: binomial, negative binomial, Poisson, gamma, chi-square, beta, normal, multivariate normal, $t$, and $F$ distributions. Key formulas: (3.1.2) binomial pmf, (3.2.1) Poisson pmf, (3.3.1) gamma pdf, (3.4.1) normal pdf. Important relationships: binomial $\to$ Poisson (large $n$, small $p$), chi-square is special gamma, $t$ and $F$ involve normal and chi-square.}
